---
title: "RNA-seq Analysis Report"
subtitle: "Generated by the rnaseq-flow pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    page-layout: full
    fig-align: center
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", dpi = 96)

# plotly + DT provide the interactive plots and tables; ggplot2/dplyr/tidyr the
# data wrangling; jsonlite parses the MultiQC export. All are provisioned in the
# QUARTO_REPORT conda environment.
library(ggplot2)
library(dplyr)
library(tidyr)
library(jsonlite)
library(plotly)
library(DT)

# ---- Locate result directories --------------------------------------------
# The QUARTO_REPORT process stages the upstream result directories next to this
# .qmd, so each is found by its plain name. Any that were not produced (e.g. no
# differential expression was run) are simply absent.
deseq2_dir    <- "deseq2_output"
edger_dir     <- "edger_output"
gsea_dir      <- "gsea_output"
gprofiler_dir <- "gprofiler_output"

list_csvs <- function(d, pat) {
  if (dir.exists(d)) list.files(d, pattern = pat, full.names = TRUE) else character(0)
}

deseq2_files    <- list_csvs(deseq2_dir,    "^deseq2_results_.*\\.csv$")
edger_files     <- list_csvs(edger_dir,     "^edger_results_.*\\.csv$")
gsea_files      <- list_csvs(gsea_dir,      "^gsea_stats_.*\\.csv$")
gprofiler_files <- list_csvs(gprofiler_dir, "^gprofiler_(UP|DOWN)_.*\\.csv$")

have_deseq2    <- length(deseq2_files)    > 0
have_edger     <- length(edger_files)     > 0
have_gsea      <- length(gsea_files)      > 0
have_gprofiler <- length(gprofiler_files) > 0

mqc_hits <- list.files(pattern = "multiqc_data.json", recursive = TRUE,
                       full.names = TRUE)
mqc_json <- if (length(mqc_hits)) mqc_hits[1] else NA_character_
have_mqc <- !is.na(mqc_json)

# ---- Significance thresholds (shared throughout the report) ----------------
PADJ_CUT       <- 0.05   # adjusted-p / FDR cutoff
LFC_CUT        <- 1      # |log2 fold change| cutoff (a 2-fold change)
MAX_TABLE_ROWS <- 1000   # cap on rows shown in an interactive results table
MAX_NS_POINTS  <- 8000   # cap on non-significant points drawn per volcano

# ---- Helpers ---------------------------------------------------------------
# Read every per-contrast DE CSV into one long data frame; the contrast label
# is recovered from the file name (DE scripts name files <prefix><contrast>.csv).
read_de <- function(files, prefix) {
  if (!length(files)) return(NULL)
  bind_rows(lapply(files, function(f) {
    d <- read.csv(f, row.names = 1, stringsAsFactors = FALSE)
    d$gene_id  <- rownames(d)
    d$contrast <- sub("\\.csv$", "", sub(prefix, "", basename(f)))
    d
  }))
}

# An interactive table with consistent defaults.
nice_table <- function(df, caption = NULL) {
  DT::datatable(df, rownames = FALSE, filter = "top", caption = caption,
                options = list(pageLength = 10, scrollX = TRUE,
                               lengthMenu = c(10, 25, 50, 100)))
}
```

# Overview

This report is produced automatically at the end of every **rnaseq-flow** run.
It gathers the run's quality-control metrics and — when differential expression
was run — the differential-expression and functional-enrichment results into a
single interactive document. The plots are [plotly](https://plotly.com/r/)
widgets (hover for gene identities, zoom, pan) and the tables are searchable and
sortable. Use the code-folding toggle to reveal how any panel was built.

```{r overview-note, results="asis"}
present <- c(
  if (have_mqc)       "quality-control metrics",
  if (have_deseq2)    "DESeq2 differential expression",
  if (have_edger)     "edgeR differential expression",
  if (have_gsea)      "GSEA enrichment",
  if (have_gprofiler) "gProfiler over-representation"
)
if (length(present)) {
  cat("\n**This run produced:** ", paste(present, collapse = ", "), ".\n", sep = "")
} else {
  cat("\n*No recognised result files were found for this run.*\n")
}
```

# Quality control

The [MultiQC report](../multiqc/multiqc_report.html) is the authoritative,
per-sample QC document; the table below is a compact summary of its general
statistics for a quick scan.

```{r qc-load, include=FALSE}
stats_df <- NULL
if (have_mqc) {
  stats_df <- tryCatch({
    mqc <- fromJSON(mqc_json)
    bind_rows(mqc$report_general_stats_data, .id = "Sample")
  }, error = function(e) NULL)
}
qc_ok <- !is.null(stats_df) && nrow(stats_df) > 0
```

```{r qc-note, results="asis"}
if (!qc_ok) {
  cat("\n*No MultiQC general-statistics table was available for this run;",
      "see the MultiQC report linked above for full QC.*\n")
}
```

```{r qc-table, eval=qc_ok}
nice_table(stats_df, "MultiQC general statistics — one row per sample.")
```

# Differential expression

```{r de-note, results="asis"}
if (!have_deseq2 && !have_edger) {
  cat("\n*Differential expression was not run for this dataset. It requires a",
      "CSV samplesheet with at least two conditions and two replicates per",
      "condition, and is skipped by `--stop_at preQC/postQC`.*\n")
}
```

## DESeq2

DESeq2 log2 fold changes are shrunk with the apeglm estimator, so the volcano
panels and results table below show shrunken effect sizes; the p-values and FDR
are from the unshrunken Wald test.

```{r deseq2-note, results="asis"}
if (!have_deseq2) cat("\n*No DESeq2 results were produced in this run.*\n")
```

```{r deseq2-load, eval=have_deseq2, include=FALSE}
deseq2 <- read_de(deseq2_files, "deseq2_results_")
deseq2$significant <- with(deseq2,
  !is.na(padj) & padj < PADJ_CUT & abs(log2FoldChange) > LFC_CUT)
deseq2$direction <- with(deseq2, ifelse(significant,
  ifelse(log2FoldChange > 0, "Up", "Down"), "Not significant"))
```

### Significant genes per contrast

```{r deseq2-summary, eval=have_deseq2}
deseq2 %>%
  group_by(Contrast = contrast) %>%
  summarise(`Genes tested` = sum(!is.na(padj)),
            Up             = sum(significant & log2FoldChange > 0),
            Down           = sum(significant & log2FoldChange < 0),
            `Significant`  = sum(significant),
            .groups = "drop") %>%
  knitr::kable(caption = sprintf(
    "A gene is significant at padj < %.2f and |log2FC| > %g.", PADJ_CUT, LFC_CUT))
```

### Volcano plots

Each point is a gene; significant genes are coloured (up in red, down in blue).
Hover for the gene symbol, fold change and adjusted p-value.

```{r deseq2-volcano, eval=have_deseq2, fig.width=9, fig.height=6}
vd  <- deseq2 %>% filter(!is.na(pvalue), !is.na(log2FoldChange))
ns  <- vd %>% filter(!significant)
sig <- vd %>% filter(significant)
# Keep every significant gene; downsample the dense non-significant cloud so the
# embedded widget stays light.
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
vd <- bind_rows(sig, ns)
vd$direction <- factor(vd$direction, levels = c("Up", "Down", "Not significant"))

p <- ggplot(vd, aes(log2FoldChange, -log10(pvalue), colour = direction,
                    text = paste0(gene_name, " (", gene_id, ")",
                                  "<br>log2FC: ", round(log2FoldChange, 2),
                                  "<br>padj: ", signif(padj, 3)))) +
  geom_point(size = 0.7, alpha = 0.6) +
  facet_wrap(~ contrast, ncol = 2) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "log2 fold change", y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Sample similarity and top variable genes

```{r deseq2-images, eval=have_deseq2, out.width="80%"}
imgs <- c(file.path(deseq2_dir, "pca_plot.png"),
          file.path(deseq2_dir, "heatmap_top_var.png"))
imgs <- imgs[file.exists(imgs)]
if (length(imgs)) knitr::include_graphics(imgs)
```

The PCA plot shows sample clustering — replicates of a condition should group
together. The heatmap shows the 20 most variable genes across all samples.

### Results table

```{r deseq2-table, eval=have_deseq2}
tbl <- deseq2 %>%
  filter(significant) %>%
  arrange(padj) %>%
  transmute(contrast, gene_id, gene_name, gene_biotype,
            baseMean       = round(baseMean, 1),
            log2FoldChange = round(log2FoldChange, 3),
            pvalue, padj)
truncated <- nrow(tbl) > MAX_TABLE_ROWS
if (truncated) tbl <- head(tbl, MAX_TABLE_ROWS)
cap <- if (truncated)
  sprintf("Top %d significant genes by padj (of %d).", MAX_TABLE_ROWS,
          sum(deseq2$significant))
else
  sprintf("All %d significant genes (padj < %.2f, |log2FC| > %g).",
          nrow(tbl), PADJ_CUT, LFC_CUT)
dt <- DT::datatable(tbl, rownames = FALSE, filter = "top", caption = cap,
                    options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, c("pvalue", "padj"), digits = 3)
```

## edgeR

edgeR runs in parallel with DESeq2 as an independent cross-check, on the same
gene counts. Genes flagged by both methods are the most robust hits.

```{r edger-note, results="asis"}
if (!have_edger) cat("\n*No edgeR results were produced in this run.*\n")
```

```{r edger-load, eval=have_edger, include=FALSE}
edger <- read_de(edger_files, "edger_results_")
edger$significant <- with(edger,
  !is.na(FDR) & FDR < PADJ_CUT & abs(logFC) > LFC_CUT)
edger$direction <- with(edger, ifelse(significant,
  ifelse(logFC > 0, "Up", "Down"), "Not significant"))
```

### Significant genes per contrast

```{r edger-summary, eval=have_edger}
edger %>%
  group_by(Contrast = contrast) %>%
  summarise(`Genes tested` = sum(!is.na(FDR)),
            Up             = sum(significant & logFC > 0),
            Down           = sum(significant & logFC < 0),
            `Significant`  = sum(significant),
            .groups = "drop") %>%
  knitr::kable(caption = sprintf(
    "A gene is significant at FDR < %.2f and |logFC| > %g.", PADJ_CUT, LFC_CUT))
```

### Volcano plots

```{r edger-volcano, eval=have_edger, fig.width=9, fig.height=6}
vd  <- edger %>% filter(!is.na(PValue), !is.na(logFC))
ns  <- vd %>% filter(!significant)
sig <- vd %>% filter(significant)
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
vd <- bind_rows(sig, ns)
vd$direction <- factor(vd$direction, levels = c("Up", "Down", "Not significant"))

p <- ggplot(vd, aes(logFC, -log10(PValue), colour = direction,
                    text = paste0(gene_name, " (", gene_id, ")",
                                  "<br>logFC: ", round(logFC, 2),
                                  "<br>FDR: ", signif(FDR, 3)))) +
  geom_point(size = 0.7, alpha = 0.6) +
  facet_wrap(~ contrast, ncol = 2) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "log2 fold change", y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Sample similarity

```{r edger-images, eval=have_edger, out.width="60%"}
mds <- file.path(edger_dir, "mds_plot.png")
if (file.exists(mds)) knitr::include_graphics(mds)
```

### Results table

```{r edger-table, eval=have_edger}
tbl <- edger %>%
  filter(significant) %>%
  arrange(FDR) %>%
  transmute(contrast, gene_id, gene_name, gene_biotype,
            logFC  = round(logFC, 3),
            logCPM = round(logCPM, 2),
            PValue, FDR)
truncated <- nrow(tbl) > MAX_TABLE_ROWS
if (truncated) tbl <- head(tbl, MAX_TABLE_ROWS)
cap <- if (truncated)
  sprintf("Top %d significant genes by FDR (of %d).", MAX_TABLE_ROWS,
          sum(edger$significant))
else
  sprintf("All %d significant genes (FDR < %.2f, |logFC| > %g).",
          nrow(tbl), PADJ_CUT, LFC_CUT)
dt <- DT::datatable(tbl, rownames = FALSE, filter = "top", caption = cap,
                    options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, c("PValue", "FDR"), digits = 3)
```

## DESeq2 vs edgeR agreement

```{r agreement-note, results="asis"}
if (!(have_deseq2 && have_edger)) {
  cat("\n*The agreement summary needs both DESeq2 and edgeR results.*\n")
}
```

```{r agreement, eval=(have_deseq2 && have_edger)}
d_sig <- deseq2 %>% filter(significant) %>% distinct(contrast, gene_id)
e_sig <- edger  %>% filter(significant) %>% distinct(contrast, gene_id)
both  <- inner_join(d_sig, e_sig, by = c("contrast", "gene_id"))

d_sig %>% count(contrast, name = "DESeq2") %>%
  full_join(e_sig %>% count(contrast, name = "edgeR"), by = "contrast") %>%
  full_join(both  %>% count(contrast, name = "Both"),  by = "contrast") %>%
  mutate(across(c(DESeq2, edgeR, Both), ~ tidyr::replace_na(., 0L))) %>%
  rename(Contrast = contrast) %>%
  knitr::kable(caption = paste("Significant-gene counts per method and the",
                               "overlap. Genes in 'Both' are the highest-",
                               "confidence calls."))
```

# Functional enrichment

```{r enrich-note, results="asis"}
if (!have_gsea && !have_gprofiler) {
  cat("\n*No functional-enrichment results were produced. GSEA needs `--gmt`;",
      "both GSEA and gProfiler are skipped by `--stop_at preQC/postQC/DE`.*\n")
}
```

## GSEA

GSEA (fgsea) is run on the whole ranked DESeq2 gene list, so it captures
coordinated shifts a hard cutoff would miss. A positive NES means the gene set
is enriched in the first condition of the contrast.

```{r gsea-note, results="asis"}
if (!have_gsea) cat("\n*No GSEA results were produced (no `--gmt` supplied).*\n")
```

```{r gsea-table, eval=have_gsea}
gsea <- bind_rows(lapply(gsea_files, function(f) {
  d <- read.csv(f, stringsAsFactors = FALSE)
  d$X <- NULL
  d$contrast <- sub("\\.csv$", "", sub("gsea_stats_", "", basename(f)))
  d
}))
keep <- intersect(c("contrast", "pathway", "NES", "ES", "pval", "padj", "size"),
                  names(gsea))
gsea <- gsea[, keep, drop = FALSE]
if ("NES" %in% names(gsea)) gsea$NES <- round(gsea$NES, 3)
if ("ES"  %in% names(gsea)) gsea$ES  <- round(gsea$ES, 3)
gsea <- gsea[order(gsea$padj), ]
dt <- DT::datatable(gsea, rownames = FALSE, filter = "top",
        caption = "fgsea results across all contrasts.",
        options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, intersect(c("pval", "padj"), names(gsea)), digits = 3)
```

```{r gsea-plots, eval=have_gsea, out.width="80%"}
gplots <- list.files(gsea_dir, "^gsea_plot_.*\\.png$", full.names = TRUE)
if (length(gplots)) knitr::include_graphics(gplots)
```

## gProfiler

gProfiler tests the thresholded significant gene lists for over-represented GO
terms and pathways, separately for the up- and down-regulated genes.

```{r gprofiler-note, results="asis"}
if (!have_gprofiler) cat("\n*No gProfiler results were produced in this run.*\n")
```

```{r gprofiler-table, eval=have_gprofiler}
wanted <- c("contrast", "direction", "source", "term_id", "term_name",
            "p_value", "term_size", "intersection_size")
gpr <- bind_rows(lapply(gprofiler_files, function(f) {
  d <- read.csv(f, stringsAsFactors = FALSE)
  base <- sub("\\.csv$", "", sub("^gprofiler_", "", basename(f)))
  d$direction <- sub("_.*$", "", base)
  d$contrast  <- sub("^(UP|DOWN)_", "", base)
  d[, intersect(wanted, names(d)), drop = FALSE]
}))
if ("p_value" %in% names(gpr)) {
  gpr$p_value <- suppressWarnings(as.numeric(gpr$p_value))
  gpr <- gpr[order(gpr$p_value), ]
}
if (nrow(gpr) > MAX_TABLE_ROWS) gpr <- head(gpr, MAX_TABLE_ROWS)
dt <- DT::datatable(gpr, rownames = FALSE, filter = "top",
        caption = "gProfiler over-representation of the up/down gene lists.",
        options = list(pageLength = 10, scrollX = TRUE))
if ("p_value" %in% names(gpr)) dt <- DT::formatSignif(dt, "p_value", digits = 3)
dt
```

# Session information

```{r sessioninfo}
sessionInfo()
```
