---
title: "RNA-seq Analysis Report"
subtitle: "Generated by the rnaseq-flow pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    page-layout: full
    fig-align: center
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", dpi = 96)

# plotly + DT provide the interactive plots and tables; ggplot2/dplyr/tidyr the
# data wrangling; jsonlite parses the MultiQC export. All are provisioned in the
# QUARTO_REPORT conda environment.
library(ggplot2)
library(dplyr)
library(tidyr)
library(jsonlite)
library(plotly)
library(DT)

# ---- Locate result directories --------------------------------------------
# The QUARTO_REPORT process stages the upstream result directories next to this
# .qmd, so each is found by its plain name. Any that were not produced (e.g. no
# differential expression was run) are simply absent.
deseq2_dir    <- "deseq2_output"
edger_dir     <- "edger_output"
gsea_dir      <- "gsea_output"
gprofiler_dir <- "gprofiler_output"

list_csvs <- function(d, pat) {
  if (dir.exists(d)) list.files(d, pattern = pat, full.names = TRUE) else character(0)
}

deseq2_files    <- list_csvs(deseq2_dir,    "^deseq2_results_.*\\.csv$")
edger_files     <- list_csvs(edger_dir,     "^edger_results_.*\\.csv$")
gsea_files      <- list_csvs(gsea_dir,      "^gsea_stats_.*\\.csv$")
gprofiler_files <- list_csvs(gprofiler_dir, "^gprofiler_(UP|DOWN)_.*\\.csv$")

have_deseq2    <- length(deseq2_files)    > 0
# Set in deseq2-load once the table is read. A results directory produced before
# the unshrunken fold change was published has no such column, so the shrinkage
# comparison is skipped rather than failing.
have_mle       <- FALSE
# Likewise for the plotted coordinates behind the PCA and MDS figures, and for
# the variance-stabilised matrix behind the heatmap.
have_pca       <- FALSE
have_mds       <- FALSE
have_heatmap   <- FALSE
have_edger     <- length(edger_files)     > 0
have_gsea      <- length(gsea_files)      > 0
have_gprofiler <- length(gprofiler_files) > 0

mqc_hits <- list.files(pattern = "multiqc_data.json", recursive = TRUE,
                       full.names = TRUE)
mqc_json <- if (length(mqc_hits)) mqc_hits[1] else NA_character_

# The general-statistics table is read from MultiQC's own flat TSV rather than
# from multiqc_data.json. The JSON nests each module's per-sample metrics as
# objects, and its exact shape has changed between MultiQC versions: parsing it
# with bind_rows() yielded a frame of nested data-frame columns, which DT cannot
# render -- every reader got a "Requested unknown parameter" popup and a table of
# meaningless values. multiqc_general_stats.txt is already one row per sample and
# is stable across versions.
gs_hits <- list.files(pattern = "multiqc_general_stats.txt", recursive = TRUE,
                      full.names = TRUE)
mqc_gs  <- if (length(gs_hits)) gs_hits[1] else NA_character_

have_mqc <- !is.na(mqc_json) || !is.na(mqc_gs)

# ---- Run manifest ----------------------------------------------------------
# Written by the workflow before this report renders (buildRunManifest() in
# workflows/rnaseq.nf) and staged beside this .qmd. It is the only route by
# which the report learns how the run was *configured* — which aligner, which
# references, which gene sets. None of that can be recovered from the result
# files, so without it a reader of this HTML alone cannot tell what was run.
#
# It records configuration, not outcome: it is built before any process runs.
# What actually got produced is established separately, from which result
# directories exist (have_deseq2 and friends above).
#
# Absent for a results directory produced before the manifest existed, in which
# case the provenance panels are skipped rather than half-filled.
manifest <- if (file.exists("run_manifest.json")) {
  tryCatch(jsonlite::fromJSON("run_manifest.json", simplifyVector = FALSE),
           error = function(e) NULL)
} else {
  NULL
}
have_manifest <- !is.null(manifest)

# Nested lookup that yields NA when any level is missing. Every field is
# optional by design: the manifest stores null where provenance could not be
# established rather than guessing, and those are rendered as "not recorded".
mf <- function(...) {
  x <- manifest
  for (k in c(...)) {
    if (is.null(x) || is.null(x[[k]])) return(NA_character_)
    x <- x[[k]]
  }
  if (length(x) == 0) return(NA_character_)
  as.character(x)[1]
}

# The manifest carries free text from the command line, so everything written
# into the HTML below is escaped.
esc <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;",  x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}

# A field that was not recorded is shown as such, never omitted and never
# inferred: a release guessed from a filename is a false record that outlives
# the run and cannot afterwards be told apart from a true one.
or_unrecorded <- function(x) {
  if (length(x) == 0 || all(is.na(x)) || !nzchar(x[1])) {
    "<em style='opacity:.6'>not recorded</em>"
  } else {
    esc(x[1])
  }
}

# ---- Significance thresholds (shared throughout the report) ----------------
PADJ_CUT       <- 0.05   # adjusted-p / FDR cutoff
LFC_CUT        <- 1      # |log2 fold change| cutoff (a 2-fold change)
MAX_TABLE_ROWS <- 1000   # cap on rows shown in an interactive results table
MAX_NS_POINTS  <- 8000   # cap on non-significant points drawn per volcano

# ---- Helpers ---------------------------------------------------------------
# Read every per-contrast DE CSV into one long data frame; the contrast label
# is recovered from the file name (DE scripts name files <prefix><contrast>.csv).
read_de <- function(files, prefix) {
  if (!length(files)) return(NULL)
  bind_rows(lapply(files, function(f) {
    d <- read.csv(f, row.names = 1, stringsAsFactors = FALSE)
    d$gene_id  <- rownames(d)
    d$contrast <- sub("\\.csv$", "", sub(prefix, "", basename(f)))
    d
  }))
}

# Figures are written as both SVG and PNG by the upstream scripts. Prefer the
# SVG: it is vector, so it stays sharp at any zoom, and for these plots (few
# points) it is no larger than the raster copy. Its text is rendered as glyph
# outlines rather than <text> elements, so it is not selectable or searchable —
# sharpness, not text fidelity, is what the SVG buys here.
# Falls back to the PNG when only that exists — an older results directory, or
# a figure whose device failed — and returns nothing when neither is present.
fig_path <- function(dir, name) {
  for (ext in c(".svg", ".png")) {
    f <- file.path(dir, paste0(name, ext))
    if (file.exists(f)) return(f)
  }
  character(0)
}

# Read a figure's data table. These live in the module's reproduce/ subfolder as
# gzipped CSVs (read.csv decompresses transparently), which is where the figure
# data moved when reproduce/ was introduced. Older results directories kept them
# uncompressed beside the figures, so both layouts are accepted.
repro_csv <- function(dir, name) {
  for (f in c(file.path(dir, "reproduce", paste0(name, ".csv.gz")),
              file.path(dir, paste0(name, ".csv")))) {
    if (file.exists(f)) {
      return(tryCatch(read.csv(f, stringsAsFactors = FALSE),
                      error = function(e) NULL))
    }
  }
  NULL
}

# An interactive table with consistent defaults.
nice_table <- function(df, caption = NULL) {
  DT::datatable(df, rownames = FALSE, filter = "top", caption = caption,
                options = list(pageLength = 10, scrollX = TRUE,
                               lengthMenu = c(10, 25, 50, 100)))
}
```

```{r title-block, results="asis"}
# Logo and pipeline link, lifted into Quarto's title block.
#
# The SVG is embedded as a base64 data URI rather than inlined as markup. Chunk
# output marked `asis` is still parsed by Pandoc, which treats it as HTML — and
# SVG is case-sensitive XML that does not survive that: `viewBox` comes back as
# `viewbox` (silently ignored by the renderer), the self-closing <line> elements
# are pushed out of an early-closed <svg> into <p> wrappers, and non-ASCII text
# in the attributes is mangled. Encoding the file wholesale leaves Pandoc a
# single <img> tag with nothing to reinterpret.
#
# Quarto renders the title block before any chunk output, so the elements are
# emitted here and moved into it on load. If the script does not run they simply
# remain where they are, at the top of the document — degraded, not broken.
logo_file <- "rnaseq-flow_logo.svg"
if (file.exists(logo_file)) {
  # xfun ships with knitr and states the MIME type explicitly, so it is tried
  # first; knitr::image_uri() infers the type and in older versions needs the
  # `mime` package, which the image does not carry. Either way a failure leaves
  # the logo out rather than aborting the render.
  logo_uri <- tryCatch(
    paste0("data:image/svg+xml;base64,", xfun::base64_encode(logo_file)),
    error = function(e) tryCatch(knitr::image_uri(logo_file),
                                 error = function(e2) NULL))
  if (!is.null(logo_uri)) {
    cat("<div class='rf-logo'><img src='", logo_uri,
        "' alt='rnaseq-flow'></div>\n", sep = "")
  }
}

if (have_manifest) {
  ver  <- mf("pipeline", "version")
  home <- mf("pipeline", "home_page")
  bits <- character(0)
  if (!is.na(ver))  bits <- c(bits, paste0("version ", esc(ver)))
  if (!is.na(home)) {
    # Linked to the repository rather than to a release tag: a run from an
    # unreleased commit would make a tag link a 404.
    bits <- c(bits, paste0("<a href='", esc(home), "'>", esc(home), "</a>"))
  }
  if (length(bits)) {
    # &middot; as an entity, not the literal character: the rendering container
    # runs in the C locale, where cat() of a non-ASCII byte emits R's <U+00B7>
    # escape into the document instead of the character.
    cat("<div class='rf-source'>", paste(bits, collapse = " &nbsp;&middot;&nbsp; "),
        "</div>\n", sep = "")
  }
}
```

```{=html}
<style>
  /* The logo leads the title block, left-aligned above the heading, rather than
     sitting in the TOC sidebar: Quarto hides .margin-sidebar below its large
     breakpoint, so a sidebar mark would vanish on the phones and tablets an
     emailed report is often opened on, and it would not print either. */
  .rf-logo { width: 240px; max-width: 62%; margin: 0 0 .9rem 0; }
  .rf-logo img { width: 100%; height: auto; display: block; }
  .rf-source { font-size: .85em; opacity: .75; margin-top: -.4rem; }
  @media (max-width: 620px) {
    .rf-logo { width: 180px; }
  }
</style>
<script>
  document.addEventListener("DOMContentLoaded", function () {
    var header = document.getElementById("title-block-header");
    if (!header) { return; }
    // The logo leads the block, so it goes in front of the <h1>; the version
    // and link line belongs under the subtitle, so it goes at the end.
    var logo = document.querySelector(".rf-logo");
    if (logo) { header.insertBefore(logo, header.firstChild); }
    var source = document.querySelector(".rf-source");
    if (source) { header.appendChild(source); }
  });
</script>
```

# Overview

This report is produced automatically at the end of every **rnaseq-flow** run.
It gathers the run's quality-control metrics and — when differential expression
was run — the differential-expression and functional-enrichment results into a
single interactive document. The plots are [plotly](https://plotly.com/r/)
widgets (hover for gene identities, zoom, pan) and the tables are searchable and
sortable. Use the code-folding toggle to reveal how any panel was built.

```{r overview-note, results="asis"}
present <- c(
  if (have_mqc)       "quality-control metrics",
  if (have_deseq2)    "DESeq2 differential expression",
  if (have_edger)     "edgeR differential expression",
  if (have_gsea)      "GSEA enrichment",
  if (have_gprofiler) "gProfiler over-representation"
)
if (length(present)) {
  cat("\n**This run produced:** ", paste(present, collapse = ", "), ".\n", sep = "")
} else {
  cat("\n*No recognised result files were found for this run.*\n")
}
```

```{r overview-provenance, eval=have_manifest, results="asis"}
# How this run was configured. A one-line summary stays visible, and the full
# record folds away so it does not push the science down the page.
#
# The annotation is described differently depending on where it came from. When
# the pipeline downloaded it, reference_metadata.json beside the files records
# the source, release and assembly authoritatively. When the user supplied it,
# the release is genuinely unknowable — an Ensembl GTF's header carries the
# assembly and genebuild date but never the release, and the trailing number in
# the filename is an annotation version, not the release. So it is reported as
# not recorded rather than inferred.
ann_origin <- mf("references", "annotation", "origin")
downloaded <- identical(ann_origin, "pipeline_download")

ref_summary <- if (downloaded) {
  paste0(esc(mf("references", "annotation", "source")), " release ",
         esc(mf("references", "annotation", "release")),
         " (assembly ", esc(mf("references", "annotation", "assembly")), ")")
} else if (!is.na(mf("references", "annotation", "file"))) {
  paste0(esc(mf("references", "annotation", "file")), " <em>(user-supplied)</em>")
} else {
  "<em style='opacity:.6'>no annotation</em>"
}

summary_bits <- character(0)
if (!is.na(mf("analysis", "aligner"))) {
  summary_bits <- c(summary_bits,
                    paste0("aligned with <strong>", esc(toupper(mf("analysis", "aligner"))),
                           "</strong>"))
} else {
  summary_bits <- c(summary_bits, "entered from a supplied count matrix")
}
summary_bits <- c(summary_bits, paste0("against ", ref_summary))
if (!is.na(mf("gene_sets", "collection"))) {
  summary_bits <- c(summary_bits,
                    paste0("gene sets <code>", esc(mf("gene_sets", "collection")), "</code>"))
}
cat("\n<p style='margin-top:.9em'><strong>This run:</strong> ",
    paste(summary_bits, collapse = ", "), ".</p>\n", sep = "")

# ---- the full record -------------------------------------------------------
row <- function(label, value) {
  paste0("<tr><th style='text-align:left;padding:.25em .9em .25em 0;",
         "font-weight:600;white-space:nowrap;vertical-align:top'>", label,
         "</th><td style='padding:.25em 0;vertical-align:top'>", value, "</td></tr>")
}

rows <- c(
  row("Pipeline",      paste0(or_unrecorded(mf("pipeline", "name")), " ",
                              or_unrecorded(mf("pipeline", "version")))),
  row("Run name",      or_unrecorded(mf("pipeline", "run_name"))),
  row("Started",       or_unrecorded(mf("pipeline", "started_at"))),
  row("Nextflow",      or_unrecorded(mf("pipeline", "nextflow_version"))),
  row("Profile",       or_unrecorded(mf("pipeline", "profile"))),
  row("Entry point",   or_unrecorded(mf("analysis", "entry"))),
  row("Aligner",       or_unrecorded(mf("analysis", "aligner"))),
  row("Quantification", or_unrecorded(mf("analysis", "quantification"))),
  row("Strandedness",  or_unrecorded(mf("analysis", "strandedness"))),
  row("Design",        paste0("<code>", or_unrecorded(mf("analysis", "design")), "</code>")),
  row("Organism",      or_unrecorded(mf("analysis", "organism")))
)

# Reference rows: the downloaded case can state release and provenance; the
# user-supplied case states what is verifiable and says so.
rows <- c(rows,
  row("Annotation",    or_unrecorded(mf("references", "annotation", "file"))),
  row("Reference source",
      if (downloaded) {
        paste0(or_unrecorded(mf("references", "annotation", "source")),
               " <em>(downloaded by this pipeline)</em>")
      } else {
        "<em>user-supplied</em>"
      }),
  row("Assembly",      or_unrecorded(mf("references", "annotation", "assembly"))),
  row("Release",       or_unrecorded(mf("references", "annotation", "release"))),
  row("Annotation version", or_unrecorded(mf("references", "annotation", "annotation_version"))),
  row("Reference obtained", or_unrecorded(mf("references", "annotation", "downloaded_at")))
)

if (!downloaded && !is.na(mf("references", "annotation", "note"))) {
  rows <- c(rows, row("Reference note",
                      paste0("<em style='opacity:.75'>",
                             esc(mf("references", "annotation", "note")), "</em>")))
}

rows <- c(rows,
  row("Index",         or_unrecorded(mf("references", "index"))),
  row("Gene sets",
      if (!is.na(mf("gene_sets", "collection"))) {
        paste0(esc(mf("gene_sets", "collection")),
               " <em style='opacity:.75'>(collection read from the filename; GMT",
               " files carry no version)</em>")
      } else {
        or_unrecorded(NA_character_)
      })
)

optional <- manifest$analysis$optional
if (length(optional)) {
  rows <- c(rows, row("Optional analyses",
                      esc(paste(unlist(optional), collapse = ", "))))
}

if (!is.na(mf("pipeline", "command_line"))) {
  rows <- c(rows, row("Command",
                      paste0("<code style='word-break:break-all;font-size:.85em'>",
                             esc(mf("pipeline", "command_line")), "</code>")))
}

cat("<details style='margin:.4em 0 1.2em'>",
    "<summary style='cursor:pointer'>Full run details</summary>",
    "<table style='margin-top:.6em;border-collapse:collapse;font-size:.92em'>",
    paste(rows, collapse = ""),
    "</table></details>\n", sep = "")
```

```{r overview-no-manifest, eval=!have_manifest, results="asis"}
cat("\n*This results directory carries no run manifest, so the aligner,",
    "references and gene sets used are not recorded. Results produced by",
    "rnaseq-flow 1.5.1 or earlier predate it.*\n")
```

# Quality control

```{r qc-intro, results="asis"}
if (have_mqc) {
  cat("The [MultiQC report](../multiqc/multiqc_report.html) is the authoritative,",
      "per-sample QC document; the table below is a compact summary of its general",
      "statistics for a quick scan.\n")
} else {
  cat("\n*No MultiQC report was produced for this run. A `--counts` run enters at",
      "differential expression from a pre-computed count matrix, so it has no read",
      "or alignment QC to summarise.*\n")
}
```

```{r qc-load, include=FALSE}
stats_df <- NULL
if (!is.na(mqc_gs)) {
  stats_df <- tryCatch({
    d <- read.delim(mqc_gs, check.names = FALSE, stringsAsFactors = FALSE,
                    na.strings = c("", "NA"))
    # The provenance column is injected as MultiQC custom content, so its raw id
    # carries two layers of scaffolding. Name it plainly.
    names(d) <- sub(".*source_fastq$", "Source FASTQ", names(d))
    # MultiQC encodes the source module into every other column name as
    # "<Module>_mqc-generalstats-<module_key>-<metric>". Keep the module and the
    # metric, drop the machinery in between.
    names(d) <- sub("_mqc-generalstats-.*?-", ": ", names(d), perl = TRUE)
    # A sample only has values for the modules that saw it -- FastQC reports per
    # FASTQ, STAR per sample -- so entirely empty columns are dropped rather than
    # shown as a wall of NA.
    keep <- vapply(d, function(x) any(!is.na(x)), logical(1))
    d <- d[, keep, drop = FALSE]
    # Full double precision is noise in a summary table.
    num <- vapply(d, is.numeric, logical(1))
    d[num] <- lapply(d[num], function(x) signif(x, 4))
    d
  }, error = function(e) NULL)
}
qc_ok <- !is.null(stats_df) && nrow(stats_df) > 0
```

```{r qc-note, results="asis"}
if (!qc_ok && have_mqc) {
  cat("\n*No MultiQC general-statistics table was available for this run;",
      "see the MultiQC report linked above for full QC.*\n")
}
```

```{r qc-table, eval=qc_ok}
nice_table(stats_df, "MultiQC general statistics — one row per sample.")
```

# Differential expression

```{r de-note, results="asis"}
if (!have_deseq2 && !have_edger) {
  cat("\n*Differential expression was not run for this dataset. It requires a",
      "CSV samplesheet with at least two conditions and two replicates per",
      "condition, and is skipped by `--stop_at preQC/postQC`.*\n")
}
```

## DESeq2

DESeq2 log2 fold changes are shrunk with the apeglm estimator, so the volcano
panels and results table below show shrunken effect sizes; the p-values and FDR
are from the unshrunken Wald test.

```{r deseq2-note, results="asis"}
if (!have_deseq2) cat("\n*No DESeq2 results were produced in this run.*\n")
```

```{r deseq2-load, eval=have_deseq2, include=FALSE}
deseq2 <- read_de(deseq2_files, "deseq2_results_")
deseq2$significant <- with(deseq2,
  !is.na(padj) & padj < PADJ_CUT & abs(log2FoldChange) > LFC_CUT)
deseq2$direction <- with(deseq2, ifelse(significant,
  ifelse(log2FoldChange > 0, "Up", "Down"), "Not significant"))
have_mle <- "log2FoldChange_MLE" %in% names(deseq2)

pca_df   <- repro_csv(deseq2_dir, "pca_data")
have_pca <- !is.null(pca_df) && all(c("PC1", "PC2") %in% names(pca_df)) &&
            nrow(pca_df) > 0

hm_df  <- repro_csv(deseq2_dir, "heatmap_top_var")
hm_ann <- repro_csv(deseq2_dir, "heatmap_top_var_annotation")
# Needs at least a 2x2 block of samples x genes for the clustering below.
HM_META      <- c("gene_id", "gene_name", "gene_biotype")
have_heatmap <- !is.null(hm_df) && nrow(hm_df) > 1 &&
                length(setdiff(names(hm_df), HM_META)) > 1
# heatmaply draws the dendrograms; without it the heatmap falls back to a plain
# geom_tile grid, which keeps the same ordering but shows no trees.
have_heatmaply <- requireNamespace("heatmaply", quietly = TRUE)
```

### Effect-size shrinkage

`log2FoldChange` in the table below is the **apeglm-shrunken** estimate. Shrinkage
pulls fold changes toward zero where the evidence is weak — low counts, high
dispersion — so that a gene with three reads cannot outrank a well-measured one
on effect size alone. These two panels are the same contrast before and after:
the left is the raw maximum-likelihood fit, the right is what the rest of this
report uses. Most of the difference should be at the left-hand (low-count) end;
if the panels look alike, shrinkage changed little and the raw estimates were
already stable.

```{r deseq2-ma, eval=(have_deseq2 && have_mle), fig.width=9, fig.height=5}
md  <- deseq2 %>% filter(!is.na(baseMean), baseMean > 0, !is.na(log2FoldChange))
ns  <- md %>% filter(!significant)
sig <- md %>% filter(significant)
# Same treatment as the volcano: keep every significant gene, thin the dense
# non-significant cloud so the embedded widget stays light.
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]

md <- bind_rows(sig, ns) %>%
  tidyr::pivot_longer(c(log2FoldChange_MLE, log2FoldChange),
                      names_to = "estimate", values_to = "lfc") %>%
  mutate(
    estimate = factor(ifelse(estimate == "log2FoldChange_MLE",
                             "Unshrunken (MLE)", "apeglm-shrunken"),
                      levels = c("Unshrunken (MLE)", "apeglm-shrunken")),
    text = paste0(gene_name, " (", gene_id, ")",
                  "<br>mean count: ", signif(baseMean, 3),
                  "<br>log2FC: ", round(lfc, 2),
                  "<br>padj: ", signif(padj, 3)))

p <- ggplot(md, aes(baseMean, lfc, colour = direction, text = text)) +
  geom_point(size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0, colour = "#868e96", linewidth = 0.3) +
  scale_x_log10() +
  facet_grid(contrast ~ estimate) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "mean of normalised counts", y = "log2 fold change", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

```{r deseq2-ma-note, eval=(have_deseq2 && !have_mle), results="asis"}
cat("\n*These results predate the unshrunken fold-change column",
    "(`log2FoldChange_MLE`), so the shrinkage comparison cannot be drawn.",
    "Re-running DESeq2 will add it.*\n")
```

Colouring marks significance, which comes from the **unshrunken** Wald test in
both panels — shrinkage changes the effect-size estimate, not the test. That is
why a gene can move noticeably between panels without changing colour.

### Significant genes per contrast

```{r deseq2-summary, eval=have_deseq2}
deseq2 %>%
  group_by(Contrast = contrast) %>%
  summarise(`Genes tested` = sum(!is.na(padj)),
            Up             = sum(significant & log2FoldChange > 0),
            Down           = sum(significant & log2FoldChange < 0),
            `Significant`  = sum(significant),
            .groups = "drop") %>%
  knitr::kable(caption = sprintf(
    "A gene is significant at padj < %.2f and |log2FC| > %g.", PADJ_CUT, LFC_CUT))
```

### Volcano plots

Each point is a gene; significant genes are coloured (up in red, down in blue).
Hover for the gene symbol, fold change and adjusted p-value.

```{r deseq2-volcano, eval=have_deseq2, fig.width=9, fig.height=6}
vd  <- deseq2 %>% filter(!is.na(pvalue), !is.na(log2FoldChange))
ns  <- vd %>% filter(!significant)
sig <- vd %>% filter(significant)
# Keep every significant gene; downsample the dense non-significant cloud so the
# embedded widget stays light.
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
vd <- bind_rows(sig, ns)
vd$direction <- factor(vd$direction, levels = c("Up", "Down", "Not significant"))

p <- ggplot(vd, aes(log2FoldChange, -log10(pvalue), colour = direction,
                    text = paste0(gene_name, " (", gene_id, ")",
                                  "<br>log2FC: ", round(log2FoldChange, 2),
                                  "<br>padj: ", signif(padj, 3)))) +
  geom_point(size = 0.7, alpha = 0.6) +
  facet_wrap(~ contrast, ncol = 2) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "log2 fold change", y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Sample similarity and top variable genes

Replicates of a condition should group together. Distance in a PCA is only
meaningful if both axes are on the same scale, so the aspect ratio is locked —
what looks like a big separation on PC1 really is one.

```{r deseq2-pca, eval=(have_deseq2 && have_pca), fig.width=7, fig.height=6}
pv <- function(col) if (col %in% names(pca_df)) pca_df[[col]][1] else NA
lab <- function(axis, pct) if (is.na(pct)) axis else
                             paste0(axis, ": ", pct, "% variance")

pca_df$.text <- paste0(
  if ("name" %in% names(pca_df)) pca_df$name else rownames(pca_df),
  if ("condition" %in% names(pca_df)) paste0("<br>condition: ", pca_df$condition) else "",
  "<br>PC1: ", round(pca_df$PC1, 2),
  "<br>PC2: ", round(pca_df$PC2, 2))

p <- ggplot(pca_df, aes(PC1, PC2, text = .text))
if ("condition" %in% names(pca_df)) {
  p <- p + geom_point(aes(colour = condition), size = 3)
} else {
  p <- p + geom_point(size = 3)
}
p <- p +
  labs(x = lab("PC1", pv("percentVar_PC1")),
       y = lab("PC2", pv("percentVar_PC2")), colour = NULL) +
  theme_bw(base_size = 11)

# coord_fixed() does not survive the ggplot -> plotly conversion, so the equal
# aspect is set on the plotly object instead, which does hold.
ggplotly(p, tooltip = "text") %>%
  plotly::layout(yaxis = list(scaleanchor = "x", scaleratio = 1))
```

```{r deseq2-pca-note, eval=(have_deseq2 && !have_pca), results="asis"}
cat("\n*No PCA coordinates were found (`reproduce/pca_data.csv.gz`), so the",
    "static figure below is shown instead.*\n")
```

The heatmap below shows the 20 most variable genes across all samples, on the
variance-stabilised scale, with the condition of each sample as a colour bar
along the top. Rows and columns are ordered by the same clustering the published
figure uses — Euclidean distance, complete linkage, no scaling — so the layout
matches it exactly. Hover for the gene, the sample and its value.

```{r deseq2-heatmap, eval=(have_deseq2 && have_heatmap && have_heatmaply), fig.width=8, fig.height=7}
mat <- as.matrix(hm_df[, setdiff(names(hm_df), HM_META), drop = FALSE])
rownames(mat) <- if ("gene_name" %in% names(hm_df)) {
  ifelse(!is.na(hm_df$gene_name) & nzchar(hm_df$gene_name),
         hm_df$gene_name, hm_df$gene_id)
} else hm_df$gene_id

# The condition bar, aligned to the matrix columns.
side <- NULL
if (!is.null(hm_ann) && all(c("sample", "condition") %in% names(hm_ann))) {
  side <- data.frame(condition = as.character(hm_ann$condition),
                     row.names = hm_ann$sample)
  side <- side[colnames(mat), , drop = FALSE]
}

# distfun / hclustfun are stated rather than left to heatmaply's defaults, so
# the ordering matches draw_heatmap() in deseq2.R and therefore the published
# figure. scale = "none" for the same reason: these are absolute vst values.
heatmaply::heatmaply(
  mat,
  col_side_colors = side,
  distfun   = function(x) dist(x, method = "euclidean"),
  hclustfun = function(x) hclust(x, method = "complete"),
  scale     = "none",
  colors    = colorRampPalette(c("#f4f6fb", "#1f4e79"))(256),
  showticklabels = c(TRUE, TRUE),
  key.title = "vst",
  plot_method = "plotly")
```

```{r deseq2-heatmap-tiles, eval=(have_deseq2 && have_heatmap && !have_heatmaply), fig.width=8, fig.height=7}
# Fallback when heatmaply is not in the image: same data and same ordering, but
# drawn as a plain tile grid with no dendrograms.
mat <- as.matrix(hm_df[, setdiff(names(hm_df), HM_META), drop = FALSE])
# Label rows by gene symbol where the annotation supplied one, falling back to
# the id. The static figure uses ids throughout; both are in the hover.
rownames(mat) <- if ("gene_name" %in% names(hm_df)) {
  ifelse(!is.na(hm_df$gene_name) & nzchar(hm_df$gene_name),
         hm_df$gene_name, hm_df$gene_id)
} else hm_df$gene_id

# Same clustering parameters as draw_heatmap() in deseq2.R. hclust is
# deterministic, so stating them reproduces the published ordering exactly.
mat <- mat[hclust(dist(mat, method = "euclidean"), method = "complete")$order,
           hclust(dist(t(mat), method = "euclidean"), method = "complete")$order,
           drop = FALSE]

cond <- if (!is.null(hm_ann) && all(c("sample", "condition") %in% names(hm_ann))) {
  setNames(as.character(hm_ann$condition), hm_ann$sample)
} else NULL

long <- as.data.frame(as.table(mat), stringsAsFactors = FALSE)
names(long) <- c("gene", "sample", "value")
long$gene   <- factor(long$gene,   levels = rownames(mat))
long$sample <- factor(long$sample, levels = colnames(mat))
long$text <- paste0(
  long$gene,
  if (!is.null(cond)) paste0("<br>", long$sample, " (", cond[as.character(long$sample)], ")")
  else                paste0("<br>", long$sample),
  "<br>vst: ", round(long$value, 2))

# The condition is folded into the column label rather than drawn as a separate
# annotation bar: plotly would need a stacked subplot for that, which is a lot
# of machinery for one row of colour.
xlabs <- if (!is.null(cond)) {
  setNames(paste0(colnames(mat), " (", cond[colnames(mat)], ")"), colnames(mat))
} else setNames(colnames(mat), colnames(mat))

p <- ggplot(long, aes(sample, gene, fill = value, text = text)) +
  geom_tile(colour = "#ffffff", linewidth = 0.4) +
  scale_x_discrete(labels = xlabs) +
  # A sequential scale, because these are absolute variance-stabilised counts
  # with no meaningful midpoint. (The reproduce script exposes SCALE if you
  # would rather see row z-scores.)
  scale_fill_gradient(low = "#f4f6fb", high = "#1f4e79", name = "vst") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

ggplotly(p, tooltip = "text")
```

```{r deseq2-images, eval=have_deseq2, out.width="80%"}
# Static fallbacks only: each figure is drawn interactively above when the data
# behind it is available. Both copies are published regardless.
imgs <- c(if (!have_pca)     fig_path(deseq2_dir, "pca_plot"),
          if (!have_heatmap) fig_path(deseq2_dir, "heatmap_top_var"))
if (length(imgs)) knitr::include_graphics(imgs)
```

Sample labels on the PCA are in the hover rather than printed beside each point,
which is what made them overlap in the static figure.

### Results table

```{r deseq2-table, eval=have_deseq2}
tbl <- deseq2 %>%
  filter(significant) %>%
  arrange(padj) %>%
  transmute(contrast, gene_id, gene_name, gene_biotype,
            baseMean       = round(baseMean, 1),
            log2FoldChange = round(log2FoldChange, 3),
            pvalue, padj)
truncated <- nrow(tbl) > MAX_TABLE_ROWS
if (truncated) tbl <- head(tbl, MAX_TABLE_ROWS)
cap <- if (truncated) {
  sprintf("Top %d significant genes by padj (of %d).", MAX_TABLE_ROWS,
          sum(deseq2$significant))
} else {
  sprintf("All %d significant genes (padj < %.2f, |log2FC| > %g).",
          nrow(tbl), PADJ_CUT, LFC_CUT)
}
dt <- DT::datatable(tbl, rownames = FALSE, filter = "top", caption = cap,
                    options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, c("pvalue", "padj"), digits = 3)
```

## edgeR

edgeR runs in parallel with DESeq2 as an independent cross-check, on the same
gene counts. Genes flagged by both methods are the most robust hits.

```{r edger-note, results="asis"}
if (!have_edger) cat("\n*No edgeR results were produced in this run.*\n")
```

```{r edger-load, eval=have_edger, include=FALSE}
edger <- read_de(edger_files, "edger_results_")
edger$significant <- with(edger,
  !is.na(FDR) & FDR < PADJ_CUT & abs(logFC) > LFC_CUT)
edger$direction <- with(edger, ifelse(significant,
  ifelse(logFC > 0, "Up", "Down"), "Not significant"))

mds_df   <- repro_csv(edger_dir, "mds_data")
have_mds <- !is.null(mds_df) && all(c("x", "y") %in% names(mds_df)) &&
            nrow(mds_df) > 0
```

### Significant genes per contrast

```{r edger-summary, eval=have_edger}
edger %>%
  group_by(Contrast = contrast) %>%
  summarise(`Genes tested` = sum(!is.na(FDR)),
            Up             = sum(significant & logFC > 0),
            Down           = sum(significant & logFC < 0),
            `Significant`  = sum(significant),
            .groups = "drop") %>%
  knitr::kable(caption = sprintf(
    "A gene is significant at FDR < %.2f and |logFC| > %g.", PADJ_CUT, LFC_CUT))
```

### Fold change vs abundance

The interactive counterpart of the `smear_*` plots in `edger_output/`. Each
point is a gene: average abundance across all samples on the x axis, fold change
on the y. Low-abundance genes spread out to the left, which is the usual place
for large but poorly-supported fold changes.

Unlike DESeq2 above, these fold changes are **not shrunk** — edgeR's
quasi-likelihood test moderates the dispersion rather than the effect size, so
what you see is the raw estimate.

```{r edger-ma, eval=have_edger, fig.width=9, fig.height=5}
ed  <- edger %>% filter(!is.na(logCPM), !is.na(logFC))
ns  <- ed %>% filter(!significant)
sig <- ed %>% filter(significant)
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
ed  <- bind_rows(sig, ns) %>%
  mutate(text = paste0(gene_name, " (", gene_id, ")",
                       "<br>logCPM: ", round(logCPM, 2),
                       "<br>logFC: ", round(logFC, 2),
                       "<br>FDR: ", signif(FDR, 3)))

p <- ggplot(ed, aes(logCPM, logFC, colour = direction, text = text)) +
  geom_point(size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0, colour = "#868e96", linewidth = 0.3) +
  geom_hline(yintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  facet_wrap(~ contrast, ncol = 2) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "average abundance (logCPM)", y = "log2 fold change", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Volcano plots

```{r edger-volcano, eval=have_edger, fig.width=9, fig.height=6}
vd  <- edger %>% filter(!is.na(PValue), !is.na(logFC))
ns  <- vd %>% filter(!significant)
sig <- vd %>% filter(significant)
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
vd <- bind_rows(sig, ns)
vd$direction <- factor(vd$direction, levels = c("Up", "Down", "Not significant"))

p <- ggplot(vd, aes(logFC, -log10(PValue), colour = direction,
                    text = paste0(gene_name, " (", gene_id, ")",
                                  "<br>logFC: ", round(logFC, 2),
                                  "<br>FDR: ", signif(FDR, 3)))) +
  geom_point(size = 0.7, alpha = 0.6) +
  facet_wrap(~ contrast, ncol = 2) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "log2 fold change", y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Sample similarity

edgeR's MDS places samples by the leading log fold change between them — the
same question as the PCA above, asked a different way. Both axes are the same
kind of distance, so the aspect ratio is locked here too.

```{r edger-mds, eval=(have_edger && have_mds), fig.width=7, fig.height=6}
vex <- function(col) if (col %in% names(mds_df)) mds_df[[col]][1] else NA
lab <- function(dim, pct) if (is.na(pct)) paste("Leading logFC dim", dim) else
                             paste0("Leading logFC dim ", dim, ": ", pct, "%")

mds_df$.text <- paste0(
  if ("sample" %in% names(mds_df)) mds_df$sample else rownames(mds_df),
  if ("condition" %in% names(mds_df)) paste0("<br>condition: ", mds_df$condition) else "",
  "<br>dim 1: ", round(mds_df$x, 3),
  "<br>dim 2: ", round(mds_df$y, 3))

p <- ggplot(mds_df, aes(x, y, text = .text))
if ("condition" %in% names(mds_df)) {
  p <- p + geom_point(aes(colour = condition), size = 3)
} else {
  p <- p + geom_point(size = 3)
}
p <- p +
  labs(x = lab(1, vex("var_explained_dim1")),
       y = lab(2, vex("var_explained_dim2")), colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text") %>%
  plotly::layout(yaxis = list(scaleanchor = "x", scaleratio = 1))
```

```{r edger-images, eval=(have_edger && !have_mds), out.width="60%"}
# Fallback only: the interactive MDS above replaces this when the coordinates
# are available in reproduce/.
mds <- fig_path(edger_dir, "mds_plot")
if (length(mds)) knitr::include_graphics(mds)
```

### Results table

```{r edger-table, eval=have_edger}
tbl <- edger %>%
  filter(significant) %>%
  arrange(FDR) %>%
  transmute(contrast, gene_id, gene_name, gene_biotype,
            logFC  = round(logFC, 3),
            logCPM = round(logCPM, 2),
            PValue, FDR)
truncated <- nrow(tbl) > MAX_TABLE_ROWS
if (truncated) tbl <- head(tbl, MAX_TABLE_ROWS)
cap <- if (truncated) {
  sprintf("Top %d significant genes by FDR (of %d).", MAX_TABLE_ROWS,
          sum(edger$significant))
} else {
  sprintf("All %d significant genes (FDR < %.2f, |logFC| > %g).",
          nrow(tbl), PADJ_CUT, LFC_CUT)
}
dt <- DT::datatable(tbl, rownames = FALSE, filter = "top", caption = cap,
                    options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, c("PValue", "FDR"), digits = 3)
```

## DESeq2 vs edgeR agreement

```{r agreement-note, results="asis"}
if (!(have_deseq2 && have_edger)) {
  cat("\n*The agreement summary needs both DESeq2 and edgeR results.*\n")
}
```

```{r agreement, eval=(have_deseq2 && have_edger)}
d_sig <- deseq2 %>% filter(significant) %>% distinct(contrast, gene_id)
e_sig <- edger  %>% filter(significant) %>% distinct(contrast, gene_id)
both  <- inner_join(d_sig, e_sig, by = c("contrast", "gene_id"))

d_sig %>% count(contrast, name = "DESeq2") %>%
  full_join(e_sig %>% count(contrast, name = "edgeR"), by = "contrast") %>%
  full_join(both  %>% count(contrast, name = "Both"),  by = "contrast") %>%
  mutate(across(c(DESeq2, edgeR, Both), ~ tidyr::replace_na(., 0L))) %>%
  rename(Contrast = contrast) %>%
  knitr::kable(caption = paste("Significant-gene counts per method and the",
                               "overlap. Genes in 'Both' are the highest-",
                               "confidence calls."))
```

# Functional enrichment

```{r enrich-note, results="asis"}
if (!have_gsea && !have_gprofiler) {
  cat("\n*No functional-enrichment results were produced. GSEA needs `--gmt`;",
      "both GSEA and gProfiler are skipped by `--stop_at preQC/postQC/DE`.*\n")
}
```

## GSEA

GSEA (fgsea) is run on the whole ranked DESeq2 gene list, so it captures
coordinated shifts a hard cutoff would miss. A positive NES means the gene set
is enriched in the first condition of the contrast.

```{r gsea-note, results="asis"}
if (!have_gsea) cat("\n*No GSEA results were produced (no `--gmt` supplied).*\n")
```

```{r gsea-load, eval=have_gsea, include=FALSE}
gsea <- bind_rows(lapply(gsea_files, function(f) {
  d <- read.csv(f, stringsAsFactors = FALSE)
  d$X <- NULL   # older results carried an unnamed index column
  d$contrast <- sub("\\.csv$", "", sub("gsea_stats_", "", basename(f)))
  d
}))
keep <- intersect(c("contrast", "pathway", "NES", "ES", "pval", "padj", "size"),
                  names(gsea))
gsea <- gsea[, keep, drop = FALSE]
if ("NES" %in% names(gsea)) gsea$NES <- round(gsea$NES, 3)
if ("ES"  %in% names(gsea)) gsea$ES  <- round(gsea$ES, 3)
gsea <- gsea[order(gsea$padj), ]
# fgsea sets NES/padj to NA for pathways whose p-value could not be estimated,
# so the plot needs both present.
gsea_ok <- nrow(gsea) > 0 && all(c("NES", "padj", "pathway") %in% names(gsea)) &&
           any(!is.na(gsea$NES) & !is.na(gsea$padj))
```

### Top enriched gene sets

The `gsea_plot_*` figures in `gsea_output/` are fgsea's own table-with-sparklines
layout, which has no interactive equivalent. This is the readable substitute:
the most significant gene sets per contrast, ordered by enrichment score. A
positive NES means the set is enriched in the first condition of the contrast.

```{r gsea-dotplot, eval=(have_gsea && gsea_ok), fig.width=9, fig.height=6}
MAX_GSEA_SETS <- 20   # per contrast, by adjusted p-value

gs <- gsea %>%
  filter(!is.na(NES), !is.na(padj)) %>%
  group_by(contrast) %>%
  slice_min(padj, n = MAX_GSEA_SETS, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    # Long MSigDB / GO names dominate the panel otherwise.
    label = ifelse(nchar(pathway) > 55, paste0(substr(pathway, 1, 52), "..."),
                   pathway),
    label = factor(label, levels = unique(label[order(NES)])),
    text  = paste0(pathway,
                   "<br>NES: ", round(NES, 3),
                   "<br>padj: ", signif(padj, 3),
                   if ("size" %in% names(gsea)) paste0("<br>genes: ", size) else ""))

p <- ggplot(gs, aes(NES, label, colour = padj, text = text)) +
  geom_vline(xintercept = 0, colour = "#868e96", linewidth = 0.3) +
  geom_point(aes(size = if ("size" %in% names(gs)) size else 3)) +
  facet_wrap(~ contrast, ncol = 1, scales = "free_y") +
  scale_colour_gradient(low = "#c2255c", high = "#ced4da", name = "padj") +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  labs(x = "normalised enrichment score (NES)", y = NULL) +
  theme_bw(base_size = 10)

ggplotly(p, tooltip = "text")
```

```{r gsea-dotplot-note, eval=(have_gsea && !gsea_ok), results="asis"}
cat("\n*No gene sets with an estimable enrichment score were returned,",
    "so the summary plot is omitted.*\n")
```

### Results table

```{r gsea-table, eval=have_gsea}
dt <- DT::datatable(gsea, rownames = FALSE, filter = "top",
        caption = "fgsea results across all contrasts.",
        options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, intersect(c("pval", "padj"), names(gsea)), digits = 3)
```

```{r gsea-plots, eval=have_gsea, out.width="80%"}
# Prefer the SVG copies; fall back to any contrast that only has a PNG.
gplots <- list.files(gsea_dir, "^gsea_plot_.*\\.svg$", full.names = TRUE)
pngs   <- list.files(gsea_dir, "^gsea_plot_.*\\.png$", full.names = TRUE)
gplots <- c(gplots, pngs[!sub("\\.png$", "", pngs) %in% sub("\\.svg$", "", gplots)])
if (length(gplots)) knitr::include_graphics(sort(gplots))
```

## gProfiler

gProfiler tests the thresholded significant gene lists for over-represented GO
terms and pathways, separately for the up- and down-regulated genes.

```{r gprofiler-note, results="asis"}
if (!have_gprofiler) cat("\n*No gProfiler results were produced in this run.*\n")
```

```{r gprofiler-load, eval=have_gprofiler, include=FALSE}
wanted <- c("contrast", "direction", "source", "term_id", "term_name",
            "p_value", "term_size", "intersection_size")
gpr <- bind_rows(lapply(gprofiler_files, function(f) {
  d <- read.csv(f, stringsAsFactors = FALSE)
  base <- sub("\\.csv$", "", sub("^gprofiler_", "", basename(f)))
  d$direction <- sub("_.*$", "", base)
  d$contrast  <- sub("^(UP|DOWN)_", "", base)
  d[, intersect(wanted, names(d)), drop = FALSE]
}))
# The CSV is written with every column as character, so the numeric ones have
# to be restored before anything can be plotted or sorted by them.
for (nm in intersect(c("p_value", "term_size", "intersection_size"), names(gpr)))
  gpr[[nm]] <- suppressWarnings(as.numeric(gpr[[nm]]))
if ("p_value" %in% names(gpr)) gpr <- gpr[order(gpr$p_value), ]
gpr_ok <- nrow(gpr) > 0 && "p_value" %in% names(gpr)
```

### Enriched terms

Each point is an enriched term, arranged by source and coloured by it, with the
strongest enrichments highest. Hover for the term name, its identifier, and how
many of your genes it captured. This is the interactive counterpart of the
`gostplot_*` Manhattan plots written to `gprofiler_output/`.

```{r gprofiler-plot, eval=(have_gprofiler && gpr_ok), fig.width=9, fig.height=6}
# Laid out like gProfiler's own Manhattan plot: terms grouped along x by source,
# -log10 p up the y axis. The x position is a rank within each panel rather than
# anything meaningful, which is true of gostplot as well -- it exists to spread
# the terms out, so the axis is left unlabelled.
gp <- gpr %>%
  filter(!is.na(p_value)) %>%
  arrange(contrast, direction, source, p_value) %>%
  group_by(contrast, direction) %>%
  mutate(x = row_number()) %>%
  ungroup() %>%
  mutate(
    neglog10p = -log10(pmax(p_value, 1e-300)),
    panel     = paste0(contrast, " — ", direction),
    text      = paste0(term_name, " (", term_id, ")",
                       "<br>source: ", source,
                       "<br>p: ", signif(p_value, 3),
                       "<br>genes: ", intersection_size, " of ", term_size))

p <- ggplot(gp, aes(x, neglog10p, colour = source, text = text)) +
  geom_point(aes(size = intersection_size), alpha = 0.75) +
  facet_wrap(~ panel, scales = "free_x", ncol = 2) +
  scale_size_continuous(range = c(1.5, 6), guide = "none") +
  labs(x = NULL, y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

ggplotly(p, tooltip = "text")
```

```{r gprofiler-plot-note, eval=(have_gprofiler && !gpr_ok), results="asis"}
cat("\n*gProfiler returned no terms to plot for this run.*\n")
```

### Results table

```{r gprofiler-table, eval=have_gprofiler}
gpr_tbl <- if (nrow(gpr) > MAX_TABLE_ROWS) head(gpr, MAX_TABLE_ROWS) else gpr
dt <- DT::datatable(gpr_tbl, rownames = FALSE, filter = "top",
        caption = "gProfiler over-representation of the up/down gene lists.",
        options = list(pageLength = 10, scrollX = TRUE))
if ("p_value" %in% names(gpr_tbl)) dt <- DT::formatSignif(dt, "p_value", digits = 3)
dt
```

# Session information

```{r sessioninfo}
sessionInfo()
```
