---
title: "RNA-seq Analysis Report"
subtitle: "Generated by the rnaseq-flow pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    page-layout: full
    fig-align: center
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", dpi = 96)

# plotly + DT provide the interactive plots and tables; ggplot2/dplyr/tidyr the
# data wrangling; jsonlite parses the MultiQC export. All are provisioned in the
# QUARTO_REPORT conda environment.
library(ggplot2)
library(dplyr)
library(tidyr)
library(jsonlite)
library(plotly)
library(DT)

# ---- Locate result directories --------------------------------------------
# The QUARTO_REPORT process stages the upstream result directories next to this
# .qmd, so each is found by its plain name. Any that were not produced (e.g. no
# differential expression was run) are simply absent.
deseq2_dir    <- "deseq2_output"
edger_dir     <- "edger_output"
gsea_dir      <- "gsea_output"
gprofiler_dir <- "gprofiler_output"

list_csvs <- function(d, pat) {
  if (dir.exists(d)) list.files(d, pattern = pat, full.names = TRUE) else character(0)
}

deseq2_files    <- list_csvs(deseq2_dir,    "^deseq2_results_.*\\.csv$")
edger_files     <- list_csvs(edger_dir,     "^edger_results_.*\\.csv$")
gsea_files      <- list_csvs(gsea_dir,      "^gsea_stats_.*\\.csv$")
gprofiler_files <- list_csvs(gprofiler_dir, "^gprofiler_(UP|DOWN)_.*\\.csv$")

have_deseq2    <- length(deseq2_files)    > 0
# Set in deseq2-load once the table is read. A results directory produced before
# the unshrunken fold change was published has no such column, so the shrinkage
# comparison is skipped rather than failing.
have_mle       <- FALSE
# Likewise for the plotted coordinates behind the PCA and MDS figures, and for
# the variance-stabilised matrix behind the heatmap.
have_pca       <- FALSE
have_mds       <- FALSE
have_heatmap   <- FALSE
have_edger     <- length(edger_files)     > 0
have_gsea      <- length(gsea_files)      > 0
have_gprofiler <- length(gprofiler_files) > 0

mqc_hits <- list.files(pattern = "multiqc_data.json", recursive = TRUE,
                       full.names = TRUE)
mqc_json <- if (length(mqc_hits)) mqc_hits[1] else NA_character_

# The general-statistics table is read from MultiQC's own flat TSV rather than
# from multiqc_data.json. The JSON nests each module's per-sample metrics as
# objects, and its exact shape has changed between MultiQC versions: parsing it
# with bind_rows() yielded a frame of nested data-frame columns, which DT cannot
# render -- every reader got a "Requested unknown parameter" popup and a table of
# meaningless values. multiqc_general_stats.txt is already one row per sample and
# is stable across versions.
gs_hits <- list.files(pattern = "multiqc_general_stats.txt", recursive = TRUE,
                      full.names = TRUE)
mqc_gs  <- if (length(gs_hits)) gs_hits[1] else NA_character_

have_mqc <- !is.na(mqc_json) || !is.na(mqc_gs)

# The published MultiQC report is not called multiqc_report.html: MultiQC
# prefixes the filename with the `title:` from assets/multiqc_config.yml, giving
# "rnaseq-flow_multiqc_report.html". Hardcoding either name is wrong -- one is
# wrong today, the other breaks the moment the title changes.
#
# The data directory staged into this task carries the same prefix
# ("rnaseq-flow_multiqc_report_data"), so the report's name is derived from it
# and the link follows the config automatically. MultiQC's own default, with no
# title set, is the unprefixed pair multiqc_data/ and multiqc_report.html, which
# does not follow the rule and is handled separately.
mqc_marker <- if (!is.na(mqc_json)) mqc_json else mqc_gs
mqc_data_dir <- if (!is.na(mqc_marker)) basename(dirname(mqc_marker)) else NA_character_
# Only derive when the directory actually ends in "_data", which is MultiQC's
# own convention. Anything else means the directory was staged under a name of
# our choosing rather than MultiQC's, and inventing "<whatever>.html" from it
# produces a confident dead link -- which is exactly what happened when the
# QUARTO_REPORT input was pinned to the literal name "multiqc_data_dir".
# Falling back to MultiQC's default is still a guess, but a recognisable one.
mqc_report_name <- if (is.na(mqc_data_dir) ||
                       identical(mqc_data_dir, "multiqc_data") ||
                       !grepl("_data$", mqc_data_dir)) {
  "multiqc_report.html"
} else {
  paste0(sub("_data$", "", mqc_data_dir), ".html")
}

# ---- Run manifest ----------------------------------------------------------
# Written by the workflow before this report renders (buildRunManifest() in
# workflows/rnaseq.nf) and staged beside this .qmd. It is the only route by
# which the report learns how the run was *configured* — which aligner, which
# references, which gene sets. None of that can be recovered from the result
# files, so without it a reader of this HTML alone cannot tell what was run.
#
# It records configuration, not outcome: it is built before any process runs.
# What actually got produced is established separately, from which result
# directories exist (have_deseq2 and friends above).
#
# Absent for a results directory produced before the manifest existed, in which
# case the provenance panels are skipped rather than half-filled.
manifest <- if (file.exists("run_manifest.json")) {
  tryCatch(jsonlite::fromJSON("run_manifest.json", simplifyVector = FALSE),
           error = function(e) NULL)
} else {
  NULL
}
have_manifest <- !is.null(manifest)

# The design is only summarised when the run began from a samplesheet; a glob
# input has no condition column to read one from.
have_samples <- have_manifest &&
  !is.null(manifest$samples) &&
  length(manifest$samples$conditions) > 0

# Requested-but-skipped analyses. Absent on a clean run, and on a results
# directory produced before the manifest recorded them.
have_notices <- have_manifest && length(unlist(manifest$notices)) > 0

# Nested lookup that yields NA when any level is missing. Every field is
# optional by design: the manifest stores null where provenance could not be
# established rather than guessing, and those are rendered as "not recorded".
mf <- function(...) {
  x <- manifest
  for (k in c(...)) {
    if (is.null(x) || is.null(x[[k]])) return(NA_character_)
    x <- x[[k]]
  }
  if (length(x) == 0) return(NA_character_)
  as.character(x)[1]
}

# The manifest carries free text from the command line, so everything written
# into the HTML below is escaped.
esc <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;",  x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}

# A field that was not recorded is shown as such, never omitted and never
# inferred: a release guessed from a filename is a false record that outlives
# the run and cannot afterwards be told apart from a true one.
or_unrecorded <- function(x) {
  if (length(x) == 0 || all(is.na(x)) || !nzchar(x[1])) {
    "<em style='opacity:.6'>not recorded</em>"
  } else {
    esc(x[1])
  }
}

# ---- Significance thresholds (shared throughout the report) ----------------
PADJ_CUT       <- 0.05   # adjusted-p / FDR cutoff
LFC_CUT        <- 1      # |log2 fold change| cutoff (a 2-fold change)
MAX_TABLE_ROWS <- 1000   # cap on rows shown in an interactive results table
MAX_NS_POINTS  <- 8000   # cap on non-significant points drawn per volcano

# ---- Helpers ---------------------------------------------------------------
# Read every per-contrast DE CSV into one long data frame; the contrast label
# is recovered from the file name (DE scripts name files <prefix><contrast>.csv).
read_de <- function(files, prefix) {
  if (!length(files)) return(NULL)
  bind_rows(lapply(files, function(f) {
    d <- read.csv(f, row.names = 1, stringsAsFactors = FALSE)
    d$gene_id  <- rownames(d)
    d$contrast <- sub("\\.csv$", "", sub(prefix, "", basename(f)))
    d
  }))
}

# Figures are written as both SVG and PNG by the upstream scripts. Prefer the
# SVG: it is vector, so it stays sharp at any zoom, and for these plots (few
# points) it is no larger than the raster copy. Its text is rendered as glyph
# outlines rather than <text> elements, so it is not selectable or searchable —
# sharpness, not text fidelity, is what the SVG buys here.
# Falls back to the PNG when only that exists — an older results directory, or
# a figure whose device failed — and returns nothing when neither is present.
fig_path <- function(dir, name) {
  for (ext in c(".svg", ".png")) {
    f <- file.path(dir, paste0(name, ext))
    if (file.exists(f)) return(f)
  }
  character(0)
}

# Read a figure's data table. These live in the module's reproduce/ subfolder as
# gzipped CSVs (read.csv decompresses transparently), which is where the figure
# data moved when reproduce/ was introduced. Older results directories kept them
# uncompressed beside the figures, so both layouts are accepted.
repro_csv <- function(dir, name) {
  for (f in c(file.path(dir, "reproduce", paste0(name, ".csv.gz")),
              file.path(dir, paste0(name, ".csv")))) {
    if (file.exists(f)) {
      return(tryCatch(read.csv(f, stringsAsFactors = FALSE),
                      error = function(e) NULL))
    }
  }
  NULL
}

# An interactive table with consistent defaults.
nice_table <- function(df, caption = NULL) {
  DT::datatable(df, rownames = FALSE, filter = "top", caption = caption,
                options = list(pageLength = 10, scrollX = TRUE,
                               lengthMenu = c(10, 25, 50, 100)))
}
```

```{r title-block, results="asis"}
# Logo and pipeline link, lifted into Quarto's title block.
#
# The SVG is embedded as a base64 data URI rather than inlined as markup. Chunk
# output marked `asis` is still parsed by Pandoc, which treats it as HTML — and
# SVG is case-sensitive XML that does not survive that: `viewBox` comes back as
# `viewbox` (silently ignored by the renderer), the self-closing <line> elements
# are pushed out of an early-closed <svg> into <p> wrappers, and non-ASCII text
# in the attributes is mangled. Encoding the file wholesale leaves Pandoc a
# single <img> tag with nothing to reinterpret.
#
# Quarto renders the title block before any chunk output, so the elements are
# emitted here and moved into it on load. If the script does not run they simply
# remain where they are, at the top of the document — degraded, not broken.
logo_file <- "rnaseq-flow_logo.svg"
if (file.exists(logo_file)) {
  # xfun ships with knitr and states the MIME type explicitly, so it is tried
  # first; knitr::image_uri() infers the type and in older versions needs the
  # `mime` package, which the image does not carry. Either way a failure leaves
  # the logo out rather than aborting the render.
  logo_uri <- tryCatch(
    paste0("data:image/svg+xml;base64,", xfun::base64_encode(logo_file)),
    error = function(e) tryCatch(knitr::image_uri(logo_file),
                                 error = function(e2) NULL))
  if (!is.null(logo_uri)) {
    cat("<div class='rf-logo'><img src='", logo_uri,
        "' alt='rnaseq-flow'></div>\n", sep = "")
  }
}

if (have_manifest) {
  ver  <- mf("pipeline", "version")
  home <- mf("pipeline", "home_page")
  bits <- character(0)
  if (!is.na(ver))  bits <- c(bits, paste0("version ", esc(ver)))
  if (!is.na(home)) {
    # Linked to the repository rather than to a release tag: a run from an
    # unreleased commit would make a tag link a 404.
    bits <- c(bits, paste0("<a href='", esc(home), "'>", esc(home), "</a>"))
  }
  if (length(bits)) {
    # &middot; as an entity, not the literal character: the rendering container
    # runs in the C locale, where cat() of a non-ASCII byte emits R's <U+00B7>
    # escape into the document instead of the character.
    cat("<div class='rf-source'>", paste(bits, collapse = " &nbsp;&middot;&nbsp; "),
        "</div>\n", sep = "")
  }
}
```

```{=html}
<style>
  /* ---- palette ---------------------------------------------------------
     Namespaced with rf- so nothing here can collide with the Bootstrap
     variables the cosmo theme defines. */
  :root {
    --rf-bg:     #f7f8fa;
    --rf-card:   #ffffff;
    --rf-border: #e5e7eb;
    --rf-text:   #1f2937;
    --rf-muted:  #6b7280;
  }

  body {
    background-color: var(--rf-bg);
    color: var(--rf-text);
    font-family: system-ui, -apple-system, "Segoe UI", Roboto, Arial, sans-serif;
    font-size: 15px;
    line-height: 1.55;
  }

  /* ---- section cards ---------------------------------------------------
     Each top-level section sits on the grey page as its own white panel, so
     the boundary between one analysis and the next is visible rather than
     implied by a heading.

     Only .level1 gets a card. Quarto nests <section class="level2/3"> inside
     it, and styling those too would draw boxes within boxes. */
  main > section.level1,
  #quarto-document-content > section.level1,
  #title-block-header {
    background-color: var(--rf-card);
    border: 1px solid var(--rf-border);
    border-radius: 10px;
    padding: 1.4rem 1.6rem;
    margin-bottom: 1.15rem;
  }

  /* Nested sections inherit the card they sit in, never their own. */
  main section.level1 section {
    background-color: transparent;
    border: 0;
    padding: 0;
  }

  /* The heading already opens a card, so its default top margin would push it
     away from the card edge it is meant to sit against. */
  main > section.level1 > h1:first-child,
  #quarto-document-content > section.level1 > h1:first-child {
    margin-top: 0;
  }

  /* The TOC stays on the grey ground -- it is navigation, not content, and a
     card around it would compete with the panels it points at. */
  #quarto-margin-sidebar,
  .sidebar.margin-sidebar {
    background-color: transparent;
  }

  /* ---- run overview ----------------------------------------------------
     The four headline facts sit in a row that reflows to whatever fits, so
     the panel works from a phone up to a wide monitor without breakpoints. */
  .rf-facts {
    display: grid;
    grid-template-columns: repeat(auto-fit, minmax(210px, 1fr));
    gap: .75rem;
    margin: .2rem 0 1.2rem;
  }
  .rf-fact {
    border: 1px solid var(--rf-border);
    border-radius: 8px;
    padding: .6rem .8rem;
    background-color: #fcfcfd;
  }
  .rf-fact-k {
    display: block;
    font-size: .78em;
    text-transform: uppercase;
    letter-spacing: .04em;
    color: var(--rf-muted);
    margin-bottom: .15rem;
  }
  .rf-fact-v { display: block; font-weight: 600; }

  /* The detail table is a definition list in table clothing: labels are a
     fixed narrow column so values line up down the page. */
  table.rf-runtable {
    border-collapse: collapse;
    font-size: .93em;
    width: 100%;
  }
  table.rf-runtable th,
  table.rf-runtable td {
    padding: .3em .9em .3em 0;
    vertical-align: top;
    border: 0;
  }
  table.rf-runtable th {
    text-align: left;
    font-weight: 600;
    white-space: nowrap;
    width: 1%;
  }
  /* Group headers break the table into Execution / Analysis / Reference so a
     20-row list scans as three short ones. */
  table.rf-runtable tr.rf-group th {
    padding-top: 1em;
    font-size: .8em;
    text-transform: uppercase;
    letter-spacing: .04em;
    color: var(--rf-muted);
    font-weight: 700;
  }
  table.rf-runtable tr.rf-group:first-child th { padding-top: 0; }
  /* A full nextflow invocation is long and has no spaces to break on. */
  code.rf-cmd { word-break: break-all; font-size: .85em; }

  /* ---- sample design, and the file index ------------------------------- */
  table.rf-designtable,
  table.rf-filetable {
    border-collapse: collapse;
    font-size: .93em;
    width: auto;
    min-width: min(100%, 32rem);
  }
  table.rf-designtable th,
  table.rf-designtable td,
  table.rf-filetable th,
  table.rf-filetable td {
    padding: .35em 1.1em .35em 0;
    vertical-align: top;
    border: 0;
    border-bottom: 1px solid var(--rf-border);
  }
  table.rf-designtable thead th,
  table.rf-filetable thead th {
    text-align: left;
    font-size: .8em;
    text-transform: uppercase;
    letter-spacing: .04em;
    color: var(--rf-muted);
  }
  table.rf-designtable tbody tr:last-child td,
  table.rf-filetable tbody tr:last-child td { border-bottom: 0; }
  table.rf-designtable code,
  table.rf-filetable code {
    font-size: .92em;
    white-space: nowrap;
  }
  table.rf-filetable { width: 100%; }
  /* Group rows separate the index into Differential expression, QC and so on.
     They repeat the run table's treatment so the two read as one system. */
  table.rf-filetable tr.rf-group th {
    padding-top: 1.1em;
    font-size: .8em;
    text-transform: uppercase;
    letter-spacing: .04em;
    color: var(--rf-muted);
    font-weight: 700;
    border-bottom: 0;
  }
  table.rf-filetable tr.rf-group:first-child th { padding-top: 0; }
  p.rf-note {
    font-size: .9em;
    color: var(--rf-muted);
    margin-top: .8rem;
  }

  /* ---- requested-but-skipped notices ------------------------------------
     Amber rather than red: nothing failed, and colouring it as an error would
     misrepresent a run that completed. It has to be noticeable, because the
     rest of the document gives no other sign that something is missing. */
  .rf-notices {
    border: 1px solid #fcd9a4;
    border-left: 4px solid #d97706;
    background-color: #fffbeb;
    border-radius: 8px;
    padding: .9rem 1.1rem;
    margin: 1.2rem 0;
  }
  .rf-notices-head {
    font-weight: 700;
    color: #92400e;
    margin-bottom: .4rem;
  }
  .rf-notices ul { margin: 0 0 .5rem; padding-left: 1.2rem; }
  .rf-notices li { margin-bottom: .25rem; }
  .rf-notices p {
    margin: 0;
    font-size: .9em;
    color: #92400e;
  }

  /* ---- title block ----------------------------------------------------- */
  /* The logo leads the title block, left-aligned above the heading, rather than
     sitting in the TOC sidebar: Quarto hides .margin-sidebar below its large
     breakpoint, so a sidebar mark would vanish on the phones and tablets an
     emailed report is often opened on, and it would not print either. */
  .rf-logo { width: 240px; max-width: 62%; margin: 0 0 .9rem 0; }
  .rf-logo img { width: 100%; height: auto; display: block; }
  .rf-source { font-size: .85em; opacity: .75; margin-top: -.4rem; }
  @media (max-width: 620px) {
    .rf-logo { width: 180px; }
  }

  /* ---- printing --------------------------------------------------------
     Cards are a screen device. On paper the grey ground wastes toner and the
     borders fragment across page breaks, so the document reverts to plain
     flowed text. */
  @media print {
    body { background-color: #fff; }
    main > section.level1,
    #quarto-document-content > section.level1,
    #title-block-header {
      border: 0;
      padding: 0;
      margin-bottom: 1rem;
    }
  }
</style>
<script>
  document.addEventListener("DOMContentLoaded", function () {
    var header = document.getElementById("title-block-header");
    if (!header) { return; }
    // The logo leads the block, so it goes in front of the <h1>; the version
    // and link line belongs under the subtitle, so it goes at the end.
    var logo = document.querySelector(".rf-logo");
    if (logo) { header.insertBefore(logo, header.firstChild); }
    var source = document.querySelector(".rf-source");
    if (source) { header.appendChild(source); }
  });
</script>
```

# Overview

This report is produced automatically at the end of every **rnaseq-flow** run.
It gathers the run's quality-control metrics and — when differential expression
was run — the differential-expression and functional-enrichment results into a
single interactive document. The plots are [plotly](https://plotly.com/r/)
widgets (hover for gene identities, zoom, pan) and the tables are searchable and
sortable. Use the code-folding toggle to reveal how any panel was built.

```{r overview-note, results="asis"}
present <- c(
  if (have_mqc)       "quality-control metrics",
  if (have_deseq2)    "DESeq2 differential expression",
  if (have_edger)     "edgeR differential expression",
  if (have_gsea)      "GSEA enrichment",
  if (have_gprofiler) "gProfiler over-representation"
)
if (length(present)) {
  cat("\n**This run produced:** ", paste(present, collapse = ", "), ".\n", sep = "")
} else {
  cat("\n*No recognised result files were found for this run.*\n")
}
```

# Run overview

```{r run-overview-none, eval=!have_manifest, results="asis"}
cat("\n*This results directory carries no run manifest, so the aligner,",
    "references and gene sets used are not recorded. Results produced by",
    "rnaseq-flow 1.5.1 or earlier predate it.*\n")
```

```{r run-overview-facts, eval=have_manifest, results="asis"}
# The four questions a reader of this HTML alone cannot otherwise answer: what
# ran it, what went in, what kind of library, and what was actually done.
#
# The first three come from the manifest, which records *configuration*. The
# analysis level is built from have_* instead — those reflect which result
# directories exist, so it describes what the run produced rather than what it
# was asked to produce. A stage that was requested and then failed would
# otherwise be reported here as though it had worked.
level <- character(0)
if (have_deseq2 || have_edger) {
  callers <- c(if (have_deseq2) "DESeq2", if (have_edger) "edgeR")
  level <- c(level, paste0("Differential expression (", paste(callers, collapse = ", "), ")"))
}
if (have_gsea)      level <- c(level, "GSEA")
if (have_gprofiler) level <- c(level, "gProfiler")
opt <- unlist(manifest$analysis$optional)
if (length(opt)) level <- c(level, opt)
if (!length(level)) level <- "Quality control only"

fact <- function(label, value) {
  paste0("<div class='rf-fact'><span class='rf-fact-k'>", label,
         "</span><span class='rf-fact-v'>", value, "</span></div>")
}

# &mdash; as an entity, never the literal character: the rendering container runs
# in the C locale, where R escapes a non-ASCII byte to <U+2014> on output. That
# applies to anything R emits — cat(), a DT caption, a plotly label — not just
# to raw HTML, so every string leaving R has to stay ASCII.
input_desc <- if (!is.na(mf("analysis", "counts"))) {
  paste0("Count matrix &mdash; <code>", esc(basename(mf("analysis", "counts"))), "</code>")
} else if (!is.na(mf("analysis", "input"))) {
  paste0("Samplesheet &mdash; <code>", esc(basename(mf("analysis", "input"))), "</code>")
} else {
  or_unrecorded(NA_character_)
}

lib_desc <- if (!is.na(mf("analysis", "library_layout"))) {
  s <- mf("analysis", "strandedness")
  paste0(esc(mf("analysis", "library_layout")),
         if (!is.na(s)) paste0(", ", esc(s)) else "")
} else {
  or_unrecorded(NA_character_)
}

cat("<div class='rf-facts'>",
    fact("Pipeline", paste0(or_unrecorded(mf("pipeline", "name")), " ",
                            or_unrecorded(mf("pipeline", "version")))),
    fact("Input", input_desc),
    fact("Library", lib_desc),
    fact("Analysis", esc(paste(level, collapse = " + "))),
    "</div>\n", sep = "")
```

```{r run-overview-table, eval=have_manifest, results="asis"}
# The full record. The annotation is described differently depending on where it
# came from: when the pipeline downloaded it, reference_metadata.json beside the
# files records source, release and assembly authoritatively. When the user
# supplied it the release is genuinely unknowable — an Ensembl GTF's header
# carries the assembly and genebuild date but never the release, and the
# trailing number in the filename is an annotation version, not the release. So
# it is reported as not recorded rather than inferred.
downloaded <- identical(mf("references", "annotation", "origin"), "pipeline_download")

row <- function(label, value) {
  paste0("<tr><th>", label, "</th><td>", value, "</td></tr>")
}
group <- function(label) {
  paste0("<tr class='rf-group'><th colspan='2'>", label, "</th></tr>")
}

rows <- c(
  group("Execution"),
  row("Run name",  or_unrecorded(mf("pipeline", "run_name"))),
  row("Started",   or_unrecorded(mf("pipeline", "started_at"))),
  row("Nextflow",  or_unrecorded(mf("pipeline", "nextflow_version"))),
  row("Profile",   or_unrecorded(mf("pipeline", "profile"))),
  row("Output directory", paste0("<code>", or_unrecorded(mf("analysis", "outdir")), "</code>")),

  group("Analysis"),
  row("Entry point",    or_unrecorded(mf("analysis", "entry"))),
  row("Aligner",        or_unrecorded(mf("analysis", "aligner"))),
  row("Quantification", or_unrecorded(mf("analysis", "quantification"))),
  row("Strandedness",   or_unrecorded(mf("analysis", "strandedness"))),
  row("Design",         paste0("<code>", or_unrecorded(mf("analysis", "design")), "</code>")),
  row("Organism",       or_unrecorded(mf("analysis", "organism"))),

  group("Reference"),
  row("Annotation", or_unrecorded(mf("references", "annotation", "file"))),
  row("Source",
      if (downloaded) {
        paste0(or_unrecorded(mf("references", "annotation", "source")),
               " <em>(downloaded by this pipeline)</em>")
      } else {
        "<em>user-supplied</em>"
      }),
  row("Assembly",           or_unrecorded(mf("references", "annotation", "assembly"))),
  row("Release",            or_unrecorded(mf("references", "annotation", "release"))),
  row("Annotation version", or_unrecorded(mf("references", "annotation", "annotation_version"))),
  row("Obtained",           or_unrecorded(mf("references", "annotation", "downloaded_at"))),
  row("Index",              paste0("<code>", or_unrecorded(mf("references", "index")), "</code>")),
  row("Gene sets",
      if (!is.na(mf("gene_sets", "collection"))) {
        paste0(esc(mf("gene_sets", "collection")),
               " <em style='opacity:.75'>(collection read from the filename;",
               " GMT files carry no version)</em>")
      } else {
        or_unrecorded(NA_character_)
      })
)

if (!downloaded && !is.na(mf("references", "annotation", "note"))) {
  rows <- c(rows, row("Note", paste0("<em style='opacity:.75'>",
                                     esc(mf("references", "annotation", "note")), "</em>")))
}

if (!is.na(mf("pipeline", "command_line"))) {
  rows <- c(rows, group("Invocation"),
            row("Command", paste0("<code class='rf-cmd'>",
                                  esc(mf("pipeline", "command_line")), "</code>")))
}

cat("<table class='rf-runtable'>", paste(rows, collapse = ""), "</table>\n", sep = "")
```

```{r notices, eval=have_notices, results="asis"}
# Analyses that were requested and did not run.
#
# This section exists only when there is something to say -- a clean run renders
# nothing at all, not an empty box. It sits here, before the results, because it
# changes how everything below should be read: the run succeeded, so nothing
# else in this document indicates that part of what was asked for is missing.
cat("<div class='rf-notices'>",
    "<div class='rf-notices-head'>Requested but not performed</div>",
    "<ul>",
    paste(vapply(unlist(manifest$notices),
                 function(n) paste0("<li>", esc(n), "</li>"), character(1)),
          collapse = ""),
    "</ul>",
    "<p>The run completed; these steps were skipped for the reasons given, so the",
    " corresponding outputs are absent.</p>",
    "</div>\n", sep = "")
```

# Sample design

```{r sample-design-none, eval=!have_samples, results="asis"}
cat("\n*The experimental design is not recorded for this run. It is read from a",
    "samplesheet, so a run started from a file glob has none to describe.*\n")
```

```{r sample-design, eval=have_samples, results="asis"}
# Who was compared with whom. The condition order is the samplesheet's own, not
# alphabetical: the REF baseline is written first by convention and every fold
# change in this report is oriented against it, so re-sorting here would show
# the design in an order that contradicts the results.
conds <- manifest$samples$conditions
n_tot <- as.integer(mf("samples", "total"))
reps  <- vapply(conds, function(x) as.integer(x$n), integer(1))

cat("<p><strong>", n_tot, " samples</strong> across <strong>", length(conds),
    " conditions</strong>", if (mf("samples", "has_batch") == "TRUE") {
      " &mdash; a <code>batch</code> column is present"
    } else { "" }, ".</p>\n", sep = "")

design_rows <- vapply(conds, function(x) {
  paste0("<tr><td><strong>", esc(x$name), "</strong></td>",
         "<td style='text-align:right'>", x$n, "</td>",
         "<td>", paste(vapply(x$samples, function(s)
                                paste0("<code>", esc(s), "</code>"), character(1)),
                       collapse = " "), "</td></tr>")
}, character(1))

cat("<table class='rf-designtable'>",
    "<thead><tr><th>Condition</th><th style='text-align:right'>Replicates</th>",
    "<th>Samples</th></tr></thead><tbody>",
    paste(design_rows, collapse = ""),
    "</tbody></table>\n", sep = "")

# An unbalanced design is not an error -- DESeq2 and edgeR both handle it -- but
# it changes how much power each contrast has, and it is the kind of thing that
# is easy to miss when the samplesheet was assembled by hand.
notes <- character(0)
if (length(unique(reps)) > 1) {
  notes <- c(notes, paste0("The design is <strong>unbalanced</strong>: ",
                           paste(vapply(conds, function(x)
                                   paste0(esc(x$name), " n=", x$n), character(1)),
                                 collapse = ", "),
                           ". Both DESeq2 and edgeR handle this, but contrasts",
                           " involving the smaller groups carry less power."))
}
if (!is.na(mf("analysis", "design"))) {
  notes <- c(notes, paste0("Model: <code>", esc(mf("analysis", "design")), "</code>",
                           if (identical(mf("samples", "has_batch"), "TRUE") &&
                               !grepl("batch", mf("analysis", "design"))) {
                             paste0(" &mdash; note the samplesheet has a <code>batch</code>",
                                    " column that this model does not use.")
                           } else { "." }))
}
if (length(notes)) {
  cat("<p class='rf-note'>", paste(notes, collapse = "<br>"), "</p>\n", sep = "")
}
```

# Quality control

```{r qc-intro, results="asis"}
if (have_mqc) {
  # The link is relative to this report's own directory, so it resolves while
  # the results tree is intact and breaks if the HTML is sent on by itself. The
  # path is stated in the text as well, so a detached copy still says where to
  # look.
  cat("The [MultiQC report](../multiqc/", mqc_report_name, ") is the authoritative,",
      " per-sample QC document; the table below is a compact summary of its general",
      " statistics for a quick scan. If this document has been moved out of its",
      " results directory the link will not resolve; the file is",
      " `multiqc/", mqc_report_name, "`.\n", sep = "")
} else {
  cat("\n*No MultiQC report was produced for this run. A `--counts` run enters at",
      "differential expression from a pre-computed count matrix, so it has no read",
      "or alignment QC to summarise.*\n")
}
```

```{r qc-load, include=FALSE}
stats_df <- NULL
if (!is.na(mqc_gs)) {
  stats_df <- tryCatch({
    d <- read.delim(mqc_gs, check.names = FALSE, stringsAsFactors = FALSE,
                    na.strings = c("", "NA"))
    # The provenance column is injected as MultiQC custom content, so its raw id
    # carries two layers of scaffolding. Name it plainly.
    names(d) <- sub(".*source_fastq$", "Source FASTQ", names(d))
    # MultiQC encodes the source module into every other column name as
    # "<Module>_mqc-generalstats-<module_key>-<metric>". Keep the module and the
    # metric, drop the machinery in between.
    names(d) <- sub("_mqc-generalstats-.*?-", ": ", names(d), perl = TRUE)
    # A sample only has values for the modules that saw it -- FastQC reports per
    # FASTQ, STAR per sample -- so entirely empty columns are dropped rather than
    # shown as a wall of NA.
    keep <- vapply(d, function(x) any(!is.na(x)), logical(1))
    d <- d[, keep, drop = FALSE]
    # Full double precision is noise in a summary table.
    num <- vapply(d, is.numeric, logical(1))
    d[num] <- lapply(d[num], function(x) signif(x, 4))
    d
  }, error = function(e) NULL)
}
qc_ok <- !is.null(stats_df) && nrow(stats_df) > 0
```

```{r qc-note, results="asis"}
if (!qc_ok && have_mqc) {
  cat("\n*No MultiQC general-statistics table was available for this run;",
      "see the MultiQC report linked above for full QC.*\n")
}
```

```{r qc-table, eval=qc_ok}
# The caption is rendered as HTML, so the entity resolves; a literal em-dash
# would come through as <U+2014> (see the note in run-overview-facts).
nice_table(stats_df, "MultiQC general statistics &mdash; one row per sample.")
```

# Differential expression

```{r de-note, results="asis"}
if (!have_deseq2 && !have_edger) {
  cat("\n*Differential expression was not run for this dataset. It requires a",
      "CSV samplesheet with at least two conditions and two replicates per",
      "condition, and is skipped by `--stop_at preQC/postQC`.*\n")
}
```

## DESeq2

DESeq2 log2 fold changes are shrunk with the apeglm estimator, so the volcano
panels and results table below show shrunken effect sizes; the p-values and FDR
are from the unshrunken Wald test.

```{r deseq2-note, results="asis"}
if (!have_deseq2) cat("\n*No DESeq2 results were produced in this run.*\n")
```

```{r deseq2-load, eval=have_deseq2, include=FALSE}
deseq2 <- read_de(deseq2_files, "deseq2_results_")
deseq2$significant <- with(deseq2,
  !is.na(padj) & padj < PADJ_CUT & abs(log2FoldChange) > LFC_CUT)
deseq2$direction <- with(deseq2, ifelse(significant,
  ifelse(log2FoldChange > 0, "Up", "Down"), "Not significant"))
have_mle <- "log2FoldChange_MLE" %in% names(deseq2)

pca_df   <- repro_csv(deseq2_dir, "pca_data")
have_pca <- !is.null(pca_df) && all(c("PC1", "PC2") %in% names(pca_df)) &&
            nrow(pca_df) > 0

hm_df  <- repro_csv(deseq2_dir, "heatmap_top_var")
hm_ann <- repro_csv(deseq2_dir, "heatmap_top_var_annotation")
# Needs at least a 2x2 block of samples x genes for the clustering below.
HM_META      <- c("gene_id", "gene_name", "gene_biotype")
have_heatmap <- !is.null(hm_df) && nrow(hm_df) > 1 &&
                length(setdiff(names(hm_df), HM_META)) > 1
# heatmaply draws the dendrograms; without it the heatmap falls back to a plain
# geom_tile grid, which keeps the same ordering but shows no trees.
have_heatmaply <- requireNamespace("heatmaply", quietly = TRUE)
```

### Effect-size shrinkage

`log2FoldChange` in the table below is the **apeglm-shrunken** estimate. Shrinkage
pulls fold changes toward zero where the evidence is weak — low counts, high
dispersion — so that a gene with three reads cannot outrank a well-measured one
on effect size alone. These two panels are the same contrast before and after:
the left is the raw maximum-likelihood fit, the right is what the rest of this
report uses. Most of the difference should be at the left-hand (low-count) end;
if the panels look alike, shrinkage changed little and the raw estimates were
already stable.

```{r deseq2-ma, eval=(have_deseq2 && have_mle), fig.width=9, fig.height=5}
md  <- deseq2 %>% filter(!is.na(baseMean), baseMean > 0, !is.na(log2FoldChange))
ns  <- md %>% filter(!significant)
sig <- md %>% filter(significant)
# Same treatment as the volcano: keep every significant gene, thin the dense
# non-significant cloud so the embedded widget stays light.
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]

md <- bind_rows(sig, ns) %>%
  tidyr::pivot_longer(c(log2FoldChange_MLE, log2FoldChange),
                      names_to = "estimate", values_to = "lfc") %>%
  mutate(
    estimate = factor(ifelse(estimate == "log2FoldChange_MLE",
                             "Unshrunken (MLE)", "apeglm-shrunken"),
                      levels = c("Unshrunken (MLE)", "apeglm-shrunken")),
    text = paste0(gene_name, " (", gene_id, ")",
                  "<br>mean count: ", signif(baseMean, 3),
                  "<br>log2FC: ", round(lfc, 2),
                  "<br>padj: ", signif(padj, 3)))

p <- ggplot(md, aes(baseMean, lfc, colour = direction, text = text)) +
  geom_point(size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0, colour = "#868e96", linewidth = 0.3) +
  scale_x_log10() +
  facet_grid(contrast ~ estimate) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "mean of normalised counts", y = "log2 fold change", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

```{r deseq2-ma-note, eval=(have_deseq2 && !have_mle), results="asis"}
cat("\n*These results predate the unshrunken fold-change column",
    "(`log2FoldChange_MLE`), so the shrinkage comparison cannot be drawn.",
    "Re-running DESeq2 will add it.*\n")
```

Colouring marks significance, which comes from the **unshrunken** Wald test in
both panels — shrinkage changes the effect-size estimate, not the test. That is
why a gene can move noticeably between panels without changing colour.

### Significant genes per contrast

```{r deseq2-summary, eval=have_deseq2}
deseq2 %>%
  group_by(Contrast = contrast) %>%
  summarise(`Genes tested` = sum(!is.na(padj)),
            Up             = sum(significant & log2FoldChange > 0),
            Down           = sum(significant & log2FoldChange < 0),
            `Significant`  = sum(significant),
            .groups = "drop") %>%
  knitr::kable(caption = sprintf(
    "A gene is significant at padj < %.2f and |log2FC| > %g.", PADJ_CUT, LFC_CUT))
```

### Volcano plots

Each point is a gene; significant genes are coloured (up in red, down in blue).
Hover for the gene symbol, fold change and adjusted p-value.

```{r deseq2-volcano, eval=have_deseq2, fig.width=9, fig.height=6}
vd  <- deseq2 %>% filter(!is.na(pvalue), !is.na(log2FoldChange))
ns  <- vd %>% filter(!significant)
sig <- vd %>% filter(significant)
# Keep every significant gene; downsample the dense non-significant cloud so the
# embedded widget stays light.
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
vd <- bind_rows(sig, ns)
vd$direction <- factor(vd$direction, levels = c("Up", "Down", "Not significant"))

p <- ggplot(vd, aes(log2FoldChange, -log10(pvalue), colour = direction,
                    text = paste0(gene_name, " (", gene_id, ")",
                                  "<br>log2FC: ", round(log2FoldChange, 2),
                                  "<br>padj: ", signif(padj, 3)))) +
  geom_point(size = 0.7, alpha = 0.6) +
  facet_wrap(~ contrast, ncol = 2) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "log2 fold change", y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Sample similarity and top variable genes

Replicates of a condition should group together. Distance in a PCA is only
meaningful if both axes are on the same scale, so the aspect ratio is locked —
what looks like a big separation on PC1 really is one.

```{r deseq2-pca, eval=(have_deseq2 && have_pca), fig.width=7, fig.height=6}
pv <- function(col) if (col %in% names(pca_df)) pca_df[[col]][1] else NA
lab <- function(axis, pct) if (is.na(pct)) axis else
                             paste0(axis, ": ", pct, "% variance")

pca_df$.text <- paste0(
  if ("name" %in% names(pca_df)) pca_df$name else rownames(pca_df),
  if ("condition" %in% names(pca_df)) paste0("<br>condition: ", pca_df$condition) else "",
  "<br>PC1: ", round(pca_df$PC1, 2),
  "<br>PC2: ", round(pca_df$PC2, 2))

p <- ggplot(pca_df, aes(PC1, PC2, text = .text))
if ("condition" %in% names(pca_df)) {
  p <- p + geom_point(aes(colour = condition), size = 3)
} else {
  p <- p + geom_point(size = 3)
}
p <- p +
  labs(x = lab("PC1", pv("percentVar_PC1")),
       y = lab("PC2", pv("percentVar_PC2")), colour = NULL) +
  theme_bw(base_size = 11)

# coord_fixed() does not survive the ggplot -> plotly conversion, so the equal
# aspect is set on the plotly object instead, which does hold.
ggplotly(p, tooltip = "text") %>%
  plotly::layout(yaxis = list(scaleanchor = "x", scaleratio = 1))
```

```{r deseq2-pca-note, eval=(have_deseq2 && !have_pca), results="asis"}
cat("\n*No PCA coordinates were found (`reproduce/pca_data.csv.gz`), so the",
    "static figure below is shown instead.*\n")
```

The heatmap below shows the 20 most variable genes across all samples, on the
variance-stabilised scale, with the condition of each sample as a colour bar
along the top. Rows and columns are ordered by the same clustering the published
figure uses — Euclidean distance, complete linkage, no scaling — so the layout
matches it exactly. Hover for the gene, the sample and its value.

```{r deseq2-heatmap, eval=(have_deseq2 && have_heatmap && have_heatmaply), fig.width=8, fig.height=7}
mat <- as.matrix(hm_df[, setdiff(names(hm_df), HM_META), drop = FALSE])
rownames(mat) <- if ("gene_name" %in% names(hm_df)) {
  ifelse(!is.na(hm_df$gene_name) & nzchar(hm_df$gene_name),
         hm_df$gene_name, hm_df$gene_id)
} else hm_df$gene_id

# The condition bar, aligned to the matrix columns.
side <- NULL
if (!is.null(hm_ann) && all(c("sample", "condition") %in% names(hm_ann))) {
  side <- data.frame(condition = as.character(hm_ann$condition),
                     row.names = hm_ann$sample)
  side <- side[colnames(mat), , drop = FALSE]
}

# distfun / hclustfun are stated rather than left to heatmaply's defaults, so
# the ordering matches draw_heatmap() in deseq2.R and therefore the published
# figure. scale = "none" for the same reason: these are absolute vst values.
heatmaply::heatmaply(
  mat,
  col_side_colors = side,
  distfun   = function(x) dist(x, method = "euclidean"),
  hclustfun = function(x) hclust(x, method = "complete"),
  scale     = "none",
  colors    = colorRampPalette(c("#f4f6fb", "#1f4e79"))(256),
  showticklabels = c(TRUE, TRUE),
  key.title = "vst",
  plot_method = "plotly")
```

```{r deseq2-heatmap-tiles, eval=(have_deseq2 && have_heatmap && !have_heatmaply), fig.width=8, fig.height=7}
# Fallback when heatmaply is not in the image: same data and same ordering, but
# drawn as a plain tile grid with no dendrograms.
mat <- as.matrix(hm_df[, setdiff(names(hm_df), HM_META), drop = FALSE])
# Label rows by gene symbol where the annotation supplied one, falling back to
# the id. The static figure uses ids throughout; both are in the hover.
rownames(mat) <- if ("gene_name" %in% names(hm_df)) {
  ifelse(!is.na(hm_df$gene_name) & nzchar(hm_df$gene_name),
         hm_df$gene_name, hm_df$gene_id)
} else hm_df$gene_id

# Same clustering parameters as draw_heatmap() in deseq2.R. hclust is
# deterministic, so stating them reproduces the published ordering exactly.
mat <- mat[hclust(dist(mat, method = "euclidean"), method = "complete")$order,
           hclust(dist(t(mat), method = "euclidean"), method = "complete")$order,
           drop = FALSE]

cond <- if (!is.null(hm_ann) && all(c("sample", "condition") %in% names(hm_ann))) {
  setNames(as.character(hm_ann$condition), hm_ann$sample)
} else NULL

long <- as.data.frame(as.table(mat), stringsAsFactors = FALSE)
names(long) <- c("gene", "sample", "value")
long$gene   <- factor(long$gene,   levels = rownames(mat))
long$sample <- factor(long$sample, levels = colnames(mat))
long$text <- paste0(
  long$gene,
  if (!is.null(cond)) paste0("<br>", long$sample, " (", cond[as.character(long$sample)], ")")
  else                paste0("<br>", long$sample),
  "<br>vst: ", round(long$value, 2))

# The condition is folded into the column label rather than drawn as a separate
# annotation bar: plotly would need a stacked subplot for that, which is a lot
# of machinery for one row of colour.
xlabs <- if (!is.null(cond)) {
  setNames(paste0(colnames(mat), " (", cond[colnames(mat)], ")"), colnames(mat))
} else setNames(colnames(mat), colnames(mat))

p <- ggplot(long, aes(sample, gene, fill = value, text = text)) +
  geom_tile(colour = "#ffffff", linewidth = 0.4) +
  scale_x_discrete(labels = xlabs) +
  # A sequential scale, because these are absolute variance-stabilised counts
  # with no meaningful midpoint. (The reproduce script exposes SCALE if you
  # would rather see row z-scores.)
  scale_fill_gradient(low = "#f4f6fb", high = "#1f4e79", name = "vst") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

ggplotly(p, tooltip = "text")
```

```{r deseq2-images, eval=have_deseq2, out.width="80%"}
# Static fallbacks only: each figure is drawn interactively above when the data
# behind it is available. Both copies are published regardless.
imgs <- c(if (!have_pca)     fig_path(deseq2_dir, "pca_plot"),
          if (!have_heatmap) fig_path(deseq2_dir, "heatmap_top_var"))
if (length(imgs)) knitr::include_graphics(imgs)
```

Sample labels on the PCA are in the hover rather than printed beside each point,
which is what made them overlap in the static figure.

### Results table

```{r deseq2-table, eval=have_deseq2}
tbl <- deseq2 %>%
  filter(significant) %>%
  arrange(padj) %>%
  transmute(contrast, gene_id, gene_name, gene_biotype,
            baseMean       = round(baseMean, 1),
            log2FoldChange = round(log2FoldChange, 3),
            pvalue, padj)
truncated <- nrow(tbl) > MAX_TABLE_ROWS
if (truncated) tbl <- head(tbl, MAX_TABLE_ROWS)
cap <- if (truncated) {
  sprintf("Top %d significant genes by padj (of %d).", MAX_TABLE_ROWS,
          sum(deseq2$significant))
} else {
  sprintf("All %d significant genes (padj < %.2f, |log2FC| > %g).",
          nrow(tbl), PADJ_CUT, LFC_CUT)
}
dt <- DT::datatable(tbl, rownames = FALSE, filter = "top", caption = cap,
                    options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, c("pvalue", "padj"), digits = 3)
```

## edgeR

edgeR runs in parallel with DESeq2 as an independent cross-check, on the same
gene counts. Genes flagged by both methods are the most robust hits.

```{r edger-note, results="asis"}
if (!have_edger) cat("\n*No edgeR results were produced in this run.*\n")
```

```{r edger-load, eval=have_edger, include=FALSE}
edger <- read_de(edger_files, "edger_results_")
edger$significant <- with(edger,
  !is.na(FDR) & FDR < PADJ_CUT & abs(logFC) > LFC_CUT)
edger$direction <- with(edger, ifelse(significant,
  ifelse(logFC > 0, "Up", "Down"), "Not significant"))

mds_df   <- repro_csv(edger_dir, "mds_data")
have_mds <- !is.null(mds_df) && all(c("x", "y") %in% names(mds_df)) &&
            nrow(mds_df) > 0
```

### Significant genes per contrast

```{r edger-summary, eval=have_edger}
edger %>%
  group_by(Contrast = contrast) %>%
  summarise(`Genes tested` = sum(!is.na(FDR)),
            Up             = sum(significant & logFC > 0),
            Down           = sum(significant & logFC < 0),
            `Significant`  = sum(significant),
            .groups = "drop") %>%
  knitr::kable(caption = sprintf(
    "A gene is significant at FDR < %.2f and |logFC| > %g.", PADJ_CUT, LFC_CUT))
```

### Fold change vs abundance

The interactive counterpart of the `smear_*` plots in `edger_output/`. Each
point is a gene: average abundance across all samples on the x axis, fold change
on the y. Low-abundance genes spread out to the left, which is the usual place
for large but poorly-supported fold changes.

Unlike DESeq2 above, these fold changes are **not shrunk** — edgeR's
quasi-likelihood test moderates the dispersion rather than the effect size, so
what you see is the raw estimate.

```{r edger-ma, eval=have_edger, fig.width=9, fig.height=5}
ed  <- edger %>% filter(!is.na(logCPM), !is.na(logFC))
ns  <- ed %>% filter(!significant)
sig <- ed %>% filter(significant)
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
ed  <- bind_rows(sig, ns) %>%
  mutate(text = paste0(gene_name, " (", gene_id, ")",
                       "<br>logCPM: ", round(logCPM, 2),
                       "<br>logFC: ", round(logFC, 2),
                       "<br>FDR: ", signif(FDR, 3)))

p <- ggplot(ed, aes(logCPM, logFC, colour = direction, text = text)) +
  geom_point(size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0, colour = "#868e96", linewidth = 0.3) +
  geom_hline(yintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  facet_wrap(~ contrast, ncol = 2) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "average abundance (logCPM)", y = "log2 fold change", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Volcano plots

```{r edger-volcano, eval=have_edger, fig.width=9, fig.height=6}
vd  <- edger %>% filter(!is.na(PValue), !is.na(logFC))
ns  <- vd %>% filter(!significant)
sig <- vd %>% filter(significant)
if (nrow(ns) > MAX_NS_POINTS) ns <- ns[sample(nrow(ns), MAX_NS_POINTS), ]
vd <- bind_rows(sig, ns)
vd$direction <- factor(vd$direction, levels = c("Up", "Down", "Not significant"))

p <- ggplot(vd, aes(logFC, -log10(PValue), colour = direction,
                    text = paste0(gene_name, " (", gene_id, ")",
                                  "<br>logFC: ", round(logFC, 2),
                                  "<br>FDR: ", signif(FDR, 3)))) +
  geom_point(size = 0.7, alpha = 0.6) +
  facet_wrap(~ contrast, ncol = 2) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             colour = "#868e96", linewidth = 0.3) +
  scale_colour_manual(values = c("Up" = "#c2255c", "Down" = "#1c7ed6",
                                 "Not significant" = "#ced4da")) +
  labs(x = "log2 fold change", y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text")
```

### Sample similarity

edgeR's MDS places samples by the leading log fold change between them — the
same question as the PCA above, asked a different way. Both axes are the same
kind of distance, so the aspect ratio is locked here too.

```{r edger-mds, eval=(have_edger && have_mds), fig.width=7, fig.height=6}
vex <- function(col) if (col %in% names(mds_df)) mds_df[[col]][1] else NA
lab <- function(dim, pct) if (is.na(pct)) paste("Leading logFC dim", dim) else
                             paste0("Leading logFC dim ", dim, ": ", pct, "%")

mds_df$.text <- paste0(
  if ("sample" %in% names(mds_df)) mds_df$sample else rownames(mds_df),
  if ("condition" %in% names(mds_df)) paste0("<br>condition: ", mds_df$condition) else "",
  "<br>dim 1: ", round(mds_df$x, 3),
  "<br>dim 2: ", round(mds_df$y, 3))

p <- ggplot(mds_df, aes(x, y, text = .text))
if ("condition" %in% names(mds_df)) {
  p <- p + geom_point(aes(colour = condition), size = 3)
} else {
  p <- p + geom_point(size = 3)
}
p <- p +
  labs(x = lab(1, vex("var_explained_dim1")),
       y = lab(2, vex("var_explained_dim2")), colour = NULL) +
  theme_bw(base_size = 11)

ggplotly(p, tooltip = "text") %>%
  plotly::layout(yaxis = list(scaleanchor = "x", scaleratio = 1))
```

```{r edger-images, eval=(have_edger && !have_mds), out.width="60%"}
# Fallback only: the interactive MDS above replaces this when the coordinates
# are available in reproduce/.
mds <- fig_path(edger_dir, "mds_plot")
if (length(mds)) knitr::include_graphics(mds)
```

### Results table

```{r edger-table, eval=have_edger}
tbl <- edger %>%
  filter(significant) %>%
  arrange(FDR) %>%
  transmute(contrast, gene_id, gene_name, gene_biotype,
            logFC  = round(logFC, 3),
            logCPM = round(logCPM, 2),
            PValue, FDR)
truncated <- nrow(tbl) > MAX_TABLE_ROWS
if (truncated) tbl <- head(tbl, MAX_TABLE_ROWS)
cap <- if (truncated) {
  sprintf("Top %d significant genes by FDR (of %d).", MAX_TABLE_ROWS,
          sum(edger$significant))
} else {
  sprintf("All %d significant genes (FDR < %.2f, |logFC| > %g).",
          nrow(tbl), PADJ_CUT, LFC_CUT)
}
dt <- DT::datatable(tbl, rownames = FALSE, filter = "top", caption = cap,
                    options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, c("PValue", "FDR"), digits = 3)
```

## DESeq2 vs edgeR agreement

```{r agreement-note, results="asis"}
if (!(have_deseq2 && have_edger)) {
  cat("\n*The agreement summary needs both DESeq2 and edgeR results.*\n")
}
```

```{r agreement, eval=(have_deseq2 && have_edger)}
d_sig <- deseq2 %>% filter(significant) %>% distinct(contrast, gene_id)
e_sig <- edger  %>% filter(significant) %>% distinct(contrast, gene_id)
both  <- inner_join(d_sig, e_sig, by = c("contrast", "gene_id"))

d_sig %>% count(contrast, name = "DESeq2") %>%
  full_join(e_sig %>% count(contrast, name = "edgeR"), by = "contrast") %>%
  full_join(both  %>% count(contrast, name = "Both"),  by = "contrast") %>%
  mutate(across(c(DESeq2, edgeR, Both), ~ tidyr::replace_na(., 0L))) %>%
  rename(Contrast = contrast) %>%
  knitr::kable(caption = paste("Significant-gene counts per method and the",
                               "overlap. Genes in 'Both' are the highest-",
                               "confidence calls."))
```

# Functional enrichment

```{r enrich-note, results="asis"}
if (!have_gsea && !have_gprofiler) {
  cat("\n*No functional-enrichment results were produced. GSEA needs `--gmt`;",
      "both GSEA and gProfiler are skipped by `--stop_at preQC/postQC/DE`.*\n")
}
```

## GSEA

GSEA (fgsea) is run on the whole ranked DESeq2 gene list, so it captures
coordinated shifts a hard cutoff would miss. A positive NES means the gene set
is enriched in the first condition of the contrast.

```{r gsea-note, results="asis"}
if (!have_gsea) cat("\n*No GSEA results were produced (no `--gmt` supplied).*\n")
```

```{r gsea-load, eval=have_gsea, include=FALSE}
gsea <- bind_rows(lapply(gsea_files, function(f) {
  d <- read.csv(f, stringsAsFactors = FALSE)
  d$X <- NULL   # older results carried an unnamed index column
  d$contrast <- sub("\\.csv$", "", sub("gsea_stats_", "", basename(f)))
  d
}))
keep <- intersect(c("contrast", "pathway", "NES", "ES", "pval", "padj", "size"),
                  names(gsea))
gsea <- gsea[, keep, drop = FALSE]
if ("NES" %in% names(gsea)) gsea$NES <- round(gsea$NES, 3)
if ("ES"  %in% names(gsea)) gsea$ES  <- round(gsea$ES, 3)
gsea <- gsea[order(gsea$padj), ]
# fgsea sets NES/padj to NA for pathways whose p-value could not be estimated,
# so the plot needs both present.
gsea_ok <- nrow(gsea) > 0 && all(c("NES", "padj", "pathway") %in% names(gsea)) &&
           any(!is.na(gsea$NES) & !is.na(gsea$padj))
```

### Top enriched gene sets

The `gsea_plot_*` figures in `gsea_output/` are fgsea's own table-with-sparklines
layout, which has no interactive equivalent. This is the readable substitute:
the most significant gene sets per contrast, ordered by enrichment score. A
positive NES means the set is enriched in the first condition of the contrast.

```{r gsea-dotplot, eval=(have_gsea && gsea_ok), fig.width=9, fig.height=6}
MAX_GSEA_SETS <- 20   # per contrast, by adjusted p-value

gs <- gsea %>%
  filter(!is.na(NES), !is.na(padj)) %>%
  group_by(contrast) %>%
  slice_min(padj, n = MAX_GSEA_SETS, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    # Long MSigDB / GO names dominate the panel otherwise.
    label = ifelse(nchar(pathway) > 55, paste0(substr(pathway, 1, 52), "..."),
                   pathway),
    label = factor(label, levels = unique(label[order(NES)])),
    text  = paste0(pathway,
                   "<br>NES: ", round(NES, 3),
                   "<br>padj: ", signif(padj, 3),
                   if ("size" %in% names(gsea)) paste0("<br>genes: ", size) else ""))

p <- ggplot(gs, aes(NES, label, colour = padj, text = text)) +
  geom_vline(xintercept = 0, colour = "#868e96", linewidth = 0.3) +
  geom_point(aes(size = if ("size" %in% names(gs)) size else 3)) +
  facet_wrap(~ contrast, ncol = 1, scales = "free_y") +
  scale_colour_gradient(low = "#c2255c", high = "#ced4da", name = "padj") +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  labs(x = "normalised enrichment score (NES)", y = NULL) +
  theme_bw(base_size = 10)

ggplotly(p, tooltip = "text")
```

```{r gsea-dotplot-note, eval=(have_gsea && !gsea_ok), results="asis"}
cat("\n*No gene sets with an estimable enrichment score were returned,",
    "so the summary plot is omitted.*\n")
```

### Results table

```{r gsea-table, eval=have_gsea}
dt <- DT::datatable(gsea, rownames = FALSE, filter = "top",
        caption = "fgsea results across all contrasts.",
        options = list(pageLength = 10, scrollX = TRUE))
DT::formatSignif(dt, intersect(c("pval", "padj"), names(gsea)), digits = 3)
```

```{r gsea-plots, eval=have_gsea, out.width="80%"}
# Prefer the SVG copies; fall back to any contrast that only has a PNG.
gplots <- list.files(gsea_dir, "^gsea_plot_.*\\.svg$", full.names = TRUE)
pngs   <- list.files(gsea_dir, "^gsea_plot_.*\\.png$", full.names = TRUE)
gplots <- c(gplots, pngs[!sub("\\.png$", "", pngs) %in% sub("\\.svg$", "", gplots)])
if (length(gplots)) knitr::include_graphics(sort(gplots))
```

## gProfiler

gProfiler tests the thresholded significant gene lists for over-represented GO
terms and pathways, separately for the up- and down-regulated genes.

```{r gprofiler-note, results="asis"}
if (!have_gprofiler) cat("\n*No gProfiler results were produced in this run.*\n")
```

```{r gprofiler-load, eval=have_gprofiler, include=FALSE}
wanted <- c("contrast", "direction", "source", "term_id", "term_name",
            "p_value", "term_size", "intersection_size")
gpr <- bind_rows(lapply(gprofiler_files, function(f) {
  d <- read.csv(f, stringsAsFactors = FALSE)
  base <- sub("\\.csv$", "", sub("^gprofiler_", "", basename(f)))
  d$direction <- sub("_.*$", "", base)
  d$contrast  <- sub("^(UP|DOWN)_", "", base)
  d[, intersect(wanted, names(d)), drop = FALSE]
}))
# The CSV is written with every column as character, so the numeric ones have
# to be restored before anything can be plotted or sorted by them.
for (nm in intersect(c("p_value", "term_size", "intersection_size"), names(gpr)))
  gpr[[nm]] <- suppressWarnings(as.numeric(gpr[[nm]]))
if ("p_value" %in% names(gpr)) gpr <- gpr[order(gpr$p_value), ]
gpr_ok <- nrow(gpr) > 0 && "p_value" %in% names(gpr)
```

### Enriched terms

Each point is an enriched term, arranged by source and coloured by it, with the
strongest enrichments highest. Hover for the term name, its identifier, and how
many of your genes it captured. This is the interactive counterpart of the
`gostplot_*` Manhattan plots written to `gprofiler_output/`.

```{r gprofiler-plot, eval=(have_gprofiler && gpr_ok), fig.width=9, fig.height=6}
# Laid out like gProfiler's own Manhattan plot: terms grouped along x by source,
# -log10 p up the y axis. The x position is a rank within each panel rather than
# anything meaningful, which is true of gostplot as well -- it exists to spread
# the terms out, so the axis is left unlabelled.
gp <- gpr %>%
  filter(!is.na(p_value)) %>%
  arrange(contrast, direction, source, p_value) %>%
  group_by(contrast, direction) %>%
  mutate(x = row_number()) %>%
  ungroup() %>%
  mutate(
    neglog10p = -log10(pmax(p_value, 1e-300)),
    # A plain hyphen, not an entity: this is a facet label drawn inside the
    # plot as text, so "&mdash;" would appear literally rather than resolving.
    panel     = paste0(contrast, " - ", direction),
    text      = paste0(term_name, " (", term_id, ")",
                       "<br>source: ", source,
                       "<br>p: ", signif(p_value, 3),
                       "<br>genes: ", intersection_size, " of ", term_size))

p <- ggplot(gp, aes(x, neglog10p, colour = source, text = text)) +
  geom_point(aes(size = intersection_size), alpha = 0.75) +
  facet_wrap(~ panel, scales = "free_x", ncol = 2) +
  scale_size_continuous(range = c(1.5, 6), guide = "none") +
  labs(x = NULL, y = "-log10 p-value", colour = NULL) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

ggplotly(p, tooltip = "text")
```

```{r gprofiler-plot-note, eval=(have_gprofiler && !gpr_ok), results="asis"}
cat("\n*gProfiler returned no terms to plot for this run.*\n")
```

### Results table

```{r gprofiler-table, eval=have_gprofiler}
gpr_tbl <- if (nrow(gpr) > MAX_TABLE_ROWS) head(gpr, MAX_TABLE_ROWS) else gpr
dt <- DT::datatable(gpr_tbl, rownames = FALSE, filter = "top",
        caption = "gProfiler over-representation of the up/down gene lists.",
        options = list(pageLength = 10, scrollX = TRUE))
if ("p_value" %in% names(gpr_tbl)) dt <- DT::formatSignif(dt, "p_value", digits = 3)
dt
```

# Files and locations

```{r files-intro, results="asis"}
cat("Where this run wrote everything, relative to its results directory.",
    "The links resolve while this document sits in `quarto_report/` alongside",
    "the other output folders; if it has been sent on by itself they will not,",
    "so every location is given as text as well.\n")
```

```{r files-table, results="asis"}
# An index of where this run wrote its outputs, built from the route the run
# actually took: a pseudo-aligner has no BAMs to index or coverage to plot, a
# --counts run has no reads to QC, and optional analyses appear only when they
# were requested.
#
# These are locations, not a presence check. The report is given only its own
# inputs, so it cannot see most of these directories -- listing a path is
# useful regardless, and a directory that is missing is informative in itself.
aligner   <- mf("analysis", "aligner")
bam_route <- !is.na(aligner) && aligner %in% c("star", "hisat2")
tx_route  <- !is.na(aligner) && aligner %in% c("salmon", "kallisto")
optional  <- unlist(manifest$analysis$optional)

frow <- function(group, name, desc, path) {
  data.frame(group = group, name = name, desc = desc, path = path,
             stringsAsFactors = FALSE)
}

items <- list()
add <- function(...) items[[length(items) + 1]] <<- frow(...)

if (have_deseq2) {
  add("Differential expression", "DESeq2",
      "Result tables per contrast, figures, and a reproduce/ folder that redraws them",
      "deseq2_output/")
}
if (have_edger) {
  add("Differential expression", "edgeR",
      "Result tables per contrast, figures, and a reproduce/ folder", "edger_output/")
}
if (have_gsea) {
  add("Functional enrichment", "GSEA",
      "Enrichment statistics, ranks, and the gene-set plots", "gsea_output/")
}
if (have_gprofiler) {
  add("Functional enrichment", "gProfiler",
      "Over-representation of the up and down gene lists", "gprofiler_output/")
}

if (bam_route) {
  add("Quantification", "featureCounts",
      "Gene-level count matrix and per-sample assignment summaries", "featurecounts/")
} else if (tx_route) {
  add("Quantification", "tximport",
      "Transcript quantification summarised to gene level, and the txi object",
      "tximport/")
}

add("Quality control", "MultiQC",
    "The authoritative per-sample QC report, across every tool that ran",
    paste0("multiqc/", mqc_report_name))
if (!identical(mf("analysis", "entry"), "count matrix")) {
  add("Quality control", "FastQC", "Per-sample raw read quality", "fastqc/")
  add("Quality control", "fastp", "Adapter and quality trimming reports", "fastp/")
}
if (bam_route) {
  add("Quality control", "RSeQC",
      "Strandedness inference, read distribution and gene-body coverage", "rseqc/")
}

if (!is.na(aligner)) {
  add("Alignment and quantification", toupper(aligner),
      if (bam_route) "Sorted BAM files, their indexes and alignment logs"
      else "Per-sample quantification directories",
      paste0(aligner, "/"))
}
if (bam_route) {
  add("Alignment and quantification", "Coverage tracks",
      "BigWig tracks for a genome browser", "bam_to_bigwig/")
}

for (o in optional) {
  d <- if (grepl("splicing", o))        "diffsplice_output/"
       else if (grepl("transcript usage", o)) "dtu_output/"
       else if (grepl("isoform", o))    "isoform_switch_output/"
       else if (grepl("fusion", o))     "star_fusion/"
       else NA_character_
  if (!is.na(d)) add("Optional analyses", o, "Requested for this run", d)
}
if (bam_route) {
  add("Optional analyses", "rMATS",
      "Alternative splicing events, when the run included it", "rmats_output/")
}

add("Provenance", "Run manifest",
    "How this run was configured, machine-readable; the source of section 2",
    "pipeline_info/run_manifest.json")
add("Provenance", "Run summary",
    "Duration, per-process resource use and completion status, written after this report",
    "pipeline_info/run_summary.html")
add("Provenance", "Software versions",
    "Every tool and version collected at run time", "pipeline_info/software_versions_mqc.yml")
add("Provenance", "Execution trace",
    "Nextflow's own timeline, report and trace files", "pipeline_info/")

tbl <- do.call(rbind, items)

out <- character(0)
for (g in unique(tbl$group)) {
  out <- c(out, paste0("<tr class='rf-group'><th colspan='3'>", esc(g), "</th></tr>"))
  part <- tbl[tbl$group == g, ]
  for (i in seq_len(nrow(part))) {
    r <- part[i, ]
    out <- c(out, paste0(
      "<tr><td><strong>", esc(r$name), "</strong></td>",
      "<td>", esc(r$desc), "</td>",
      "<td><a href='../", r$path, "'><code>", esc(r$path), "</code></a></td></tr>"))
  }
}

cat("<table class='rf-filetable'>",
    "<thead><tr><th>Output</th><th>Contents</th><th>Location</th></tr></thead><tbody>",
    paste(out, collapse = ""), "</tbody></table>\n", sep = "")
```

# Session information

```{r sessioninfo}
sessionInfo()
```
